\documentclass[11pt]{article}

% Packages
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{cleveref}

% Custom commands
\newcommand{\dT}{\Delta T}
\newcommand{\dTstar}{\Delta T^{*}}
\newcommand{\tcal}{T_{\mathrm{cal}}^{\max}}
\newcommand{\tend}{t_{\mathrm{end}}}
\newcommand{\sigr}{\sigma_{\hat{r}}}
\newcommand{\sigrdot}{\sigma_{\dot{\hat{r}}}}
\newcommand{\sigmodel}{\sigma_{\mathrm{model}}}
\newcommand{\sigtrend}{\sigma_{\mathrm{trend}}}
\newcommand{\MSEtrend}{\mathrm{MSE}_{\mathrm{trend}}}
\newcommand{\MSEmodel}{\mathrm{MSE}_{\mathrm{model}}}

\title{Data indicate that sea level rise projections are too low by at least half}
\author{Minchew et al.}
\date{}
\begin{document}
\maketitle

\begin{abstract}
Coastal adaptation worldwide is planned around sea-level projections that rely on ice-sheet models whose outputs diverge from satellite observations within years of initialization. We construct a data-driven alternative, using Bayesian calibration of reduced physical models for individual sea-level contributors, leveraging independent observational records for each and validating the sum against satellite altimetry of total sea level. Median projected sea-level rise by 2100 relative to pre-industrial levels ranges from 1.1~m (0.8--1.7, 90\% range) under low emissions to 1.3~m (1.0--1.9) under high-but-plausible emissions---roughly double current consensus estimates---with an approximately 75\% probability of exceeding one meter across the range of plausible emission pathways. Projection uncertainty is dominated by West Antarctic ice dynamics, which is largely independent of the emission pathway and four times larger than the spread across warming scenarios. The emission-scenario spread is driven by the temperature-sensitive components---primarily glaciers and the Greenland ice sheet---whose observationally calibrated sensitivities substantially exceed those in current consensus projections. These results indicate that the world is planning for less sea-level rise than the observations support.
\end{abstract}

\item{Introduction}
\begin{itemize}
	\item Overview of SLR 
	\item Current state of SLR projections
	\item Universal forecasting principles 
	\item Failures and limitations of current SLR forecasting models 
\end{itemize}

\item{Data and Methods}
\begin{itemize}
	\item Extrapolation from Hamlington et al quadratic fits 
	\item SLR semi-empirical models -- Bayesian rate-state: quadratic and state relaxation time; emergence of quadratic from change in dominant SLR source 
	\item SLR component-wise decompositions using physics-informed, hiearchical Bayesian 
\end{itemize}

\item{Results}
\begin{itemize}

	\item Naive forecast - data extrapolation (brief bc it's been published); 
	\begin{itemize}
		\item baseline estimation
		\item  likely lower-bound due to continued warming 
	\end{itemize}
	
	\item SLR quadratic and state relaxation time; 
	\begin{itemize}
		\item emergence of quadratic from change in dominant SLR source; 
		\item unlikely to be reliable by themselves (include analysis) but 
		\item resulting forecasts likely bracket (upper bound) truth 
	\end{itemize}
	
	\item SLR component-wise decompositions
	\begin{itemize}
		\item component-wise fits; priors + posteriors 
		\item forecasting skill 
		\item Arguably most-reduced framework that faithfully represents physics 
	\end{itemize}
	
	\item WAIS uncertainty -- Consider if simply using Bamber SEJ is appropriate here bc it's easy, honest, transparent, justifiable -- can followup with a separate paper on WAIS uncertainty 
	
	\item Uncertainty budget and value of information 
	
\end{itemize}


\item{Discussion}
\begin{itemize}
	\item Discuss results for SLR projections 
	\item Forecasting skills 
	\item Comparison with IPCC results -- fall below observed trends even for warming scenarios; fail to capture basic sensitivities observed in data; component-wise comparisons all fail tests against data 
\end{itemize}

\item{Conclusions} 

\end{enumerate}

\section{Introduction}

\textbf{Overview of SLR.}
Global mean sea level has risen approximately 20~cm since 1900, with the rate of rise accelerating from $\sim\!1.4$~mm\,yr$^{-1}$ over the 20th century to $\sim\!4.5$~mm\,yr$^{-1}$ in the most recent decade \citep{Frederikse2020, Hamlington2020}. This acceleration reflects the combined effect of ocean thermal expansion and increasing mass loss from glaciers and the Greenland and Antarctic ice sheets. For each centimeter of global mean sea level rise, roughly one million additional people are exposed to the risk of coastal flooding and displacement \citep{Kulp2019}. Sea level rise is among the most consequential and irreversible impacts of climate change: once ice is lost and oceans warmed, the commitment persists for centuries to millennia regardless of subsequent emissions reductions \citep{Clark2016}.

\textbf{Current state of SLR projections.}
The Intergovernmental Panel on Climate Change (IPCC) Sixth Assessment Report (AR6) provides sea level projections based primarily on process-based ice sheet models (the Ice Sheet Model Intercomparison Project, ISMIP6; \citealp{Seroussi2020}), emulated glacier models (GlacierMIP; \citealp{Marzeion2020}), and coupled climate model output for thermal expansion. These projections form the basis for coastal adaptation planning worldwide. The AR6 medium-confidence projections for 2100 range from approximately 0.3~m (SSP1-2.6) to 0.7~m (SSP5-8.5) relative to a 1995--2014 baseline, with a ``low-likelihood, high-impact'' upper bound of 2~m that reflects deep uncertainty in ice sheet dynamics \citep{Fox-Kemper2021}. The gap between the medium-confidence projections and the acknowledged upper bound---a factor of three---reflects not measurement uncertainty but unquantified structural uncertainty in the process models that inform the central estimates.

\textbf{Universal forecasting principles.}
The broader forecasting literature, developed across economics, meteorology, and public health, has established principles that are relevant to sea level prediction but have not been applied systematically. First, forecast models should be no more complex than the information content of the calibration data can support; increasing complexity beyond this threshold degrades out-of-sample skill \citep{Green2015}. Second, forecast combinations---weighted averages of multiple independent methods---consistently outperform individual methods, including the most sophisticated among them \citep{Makridakis2020}. Third, model skill should be demonstrated through out-of-sample validation, not in-sample fit statistics. Current sea level projections rely on process models with millions of degrees of freedom and dozens of uncertain parameterizations, constrained by observational records whose information content is small relative to the demands of the models. The forecasting literature suggests that this regime is likely to produce overconfident, poorly calibrated predictions.

\textbf{Failures and limitations of current SLR forecasting models.}
Multiple lines of evidence indicate that IPCC process-model projections systematically underestimate the rate and magnitude of sea level rise. ISMIP6 ice sheet models diverge from IMBIE satellite observations almost immediately, with most models producing ice sheets that gain mass under warming rather than losing it \citep{Aschwanden2021, Fricker2025}. Under continued warming to 4$\,^\circ$C or more, observed trends in global mean sea level exceed IPCC medium-confidence projections for all emission scenarios \citep{Hamlington2025}. Known sources of low bias include: the universal use of Glen's flow law exponent $n = 3$ when observations support $n \approx 4$, which underestimates ice sheet response by 21--35\% \citep{Millstein2022, Martin2026}; the absence of calving, subglacial hydrology, and ice damage from most models \citep{Aschwanden2021}; the sensitivity of projections to calibration methodology, with different data constraints producing factor-of-two differences in projected loss rates from the same model \citep{Goldberg2026}; and incomplete uncertainty quantification that omits parametric uncertainty, which alone may exceed inter-model spread \citep{Aschwanden2019}. These are not speculative concerns---they are documented, quantified biases in the models that inform the central estimates used for coastal planning.

In this work, we construct a data-driven, physics-informed hierarchical Bayesian framework for sea level prediction that does not depend on process models and that makes its assumptions, calibration range, and extrapolation uncertainty explicit. The framework proceeds in three steps of increasing physical complexity: (1) observational extrapolation of satellite-era trends, which provides a baseline lower bound; (2) an aggregate semi-empirical model relating the rate of sea level rise to global mean surface temperature, which captures the observed acceleration but conflates physically distinct components; and (3) a component-level decomposition that fits separate rate--temperature relationships to each contributor (thermal expansion, glaciers, Greenland ice sheet, Antarctic sub-components), identifies where the aggregate acceleration originates, and produces conditional predictive distributions with a decomposed uncertainty budget. Throughout, we calibrate exclusively against observational data, use Bayesian inference to propagate parametric uncertainty, and validate against independent datasets withheld from calibration.

\section{Methods} 
\subsection{Naive forecasts}

We use Hamlington et al results ***Brent comment: needs fleshing out***

\subsection{Semi-empirical forecasts} 

\subsection{Physics-informed forecasts}
	\subsubsection{Thermosteric}

Ocean thermal expansion is the largest single contributor to observed sea level rise and the component with the longest observational record. Rather than fitting a custom model, we adopt a hybrid approach: NOAA in situ observations for the hindcast and IPCC AR6 projections for the future, supplemented by literature-derived corrections for ocean depths not covered by the primary observations.

\paragraph{Rationale.}
The IPCC AR6 thermosteric projections are constructed from a two-layer energy balance model \citep{Geoffroy2013} emulated by FaIR \citep{Smith2019}, with the ensemble filtered against observed GSAT (1850--2019) and ocean heat content (1971--2018), and ocean heat content converted to steric sea level via CMIP6-derived expansion coefficients \citep{Kuhlbrodt2012, Fox-Kemper2021}. This pipeline uses the same ODE physics that our original custom model employed, but with observational constraints applied at the emulator level by the broader climate modeling community. There is no value in us refitting this model independently: the thermosteric response to warming is the most mature component of the sea level budget, and the IPCC approach is both peer-reviewed and observationally constrained.

\paragraph{Hindcast (1950--2025).}
The observational hindcast is built from NOAA thermosteric sea level \citep{Levitus2012}:
\begin{itemize}
	\item \textbf{0--700\,m} (1955--2025, 71 annual points): the primary record, based entirely on in situ hydrographic measurements.
	\item \textbf{0--2000\,m} (2005--2025, 21 annual points): extends the depth coverage through the Argo era.
\end{itemize}
For the pre-Argo period (1955--2005), only the 0--700\,m product is available. We add two literature-derived depth corrections to approximate the full-depth thermosteric signal:
\begin{itemize}
	\item \textbf{700--2000\,m:} $+0.15 \pm 0.05$\,mm\,yr$^{-1}$ \citep{Levitus2012, Cheng2017}, applied as a cumulative correction for years before the 0--2000\,m record begins.
	\item \textbf{Below 2000\,m:} $+0.07 \pm 0.04$\,mm\,yr$^{-1}$ \citep{Purkey2010, Desbruyeres2016}, applied over the full period.
\end{itemize}
Where the 0--2000\,m product is available (2005--2025), we use it directly and add only the below-2000\,m correction. All series are rebased to the 2005 baseline.

\paragraph{NOAA structural uncertainty.}
The NOAA thermosteric product is known to be biased low relative to other observational analyses. \citet{Li2022} showed that linear vertical interpolation of sparse hydrographic profiles, as used in the NOAA objective analysis, underestimates global ocean heat content trends by approximately 14\%, primarily due to the curvature of the temperature--depth profile. Additional sources of inter-product spread include differences in XBT fall-rate corrections \citep{Cheng2014_xbt} and mapping radius \citep{Cheng2017}. We incorporate this as a structural uncertainty of 14\% of the signal magnitude (1$\sigma$), added in quadrature with the NOAA reported standard errors and the depth correction uncertainties. This inflates the thermosteric error bars to reflect the true inter-product spread without applying a bias correction to the central estimate.

\paragraph{Projections (2020--2150).}
Future thermosteric sea level is taken directly from the IPCC AR6 \texttt{oceandynamics} component projections \citep{Fox-Kemper2021, Kopp2023}. These are available as 107-quantile distributions at decadal resolution for each SSP. We generate 2000 Monte Carlo samples by drawing rank-preserving uniform quantiles $u_i \sim U(0,1)$, interpolating the quantile function at each decadal point, and then interpolating to annual resolution. Each IPCC sample is level-matched to the corresponding observation sample at the splice year (2020), ensuring continuity.

\paragraph{Transfer function for Greenland discharge.}
The Greenland discharge model requires subsurface ocean temperature as forcing. This is obtained independently from the thermosteric component: a linear transfer function $T_{\mathrm{ocean}} = \alpha\, T_{\mathrm{surface}} + \beta$ is calibrated by regressing observed EN4 subsurface temperature (200--500\,m, Greenland peripheral waters; \citealp{Good2013}) against observed Berkeley Earth surface temperature \citep{Rohde2020}. Under SSP projections, the transfer function maps projected surface temperature to subsurface ocean temperature, which then forces the discharge delay model. The thermosteric and Greenland components are thus independent in their calibration but share a common surface temperature forcing.

	\subsubsection{Glaciers}

Mountain glaciers contribute to sea level rise through mass loss driven by surface energy balance changes under warming. We model the glacier contribution using the same rate--temperature framework as the thermosteric component, but with priors that permit moderate nonlinearity to accommodate the expected response as warming exposes new ice area and accelerates frontal ablation.

The instantaneous rate of glacier mass loss is modeled as
\begin{equation}
	\dot{H}_{\mathrm{gl}}(t) = a_{\mathrm{gl}}\, T(t)^2 + b_{\mathrm{gl}}\, T(t) + c_{\mathrm{gl}},
	\label{eq:glacier_rate}
\end{equation}
where $T(t)$ is the global mean surface temperature anomaly relative to the 1995--2005 mean, $a_{\mathrm{gl}}$ captures any nonlinear acceleration in the temperature sensitivity of glacier mass loss, $b_{\mathrm{gl}}$ is the linear mass-loss sensitivity to temperature, and $c_{\mathrm{gl}}$ is a background trend representing mass loss under pre-industrial equilibrium conditions (expected to be small but nonzero, as many glaciers were already adjusting to post--Little Ice Age warming).

The cumulative glacier sea level contribution is
\begin{equation}
	H_{\mathrm{gl}}(t) = a_{\mathrm{gl}}\, I_2(t) + b_{\mathrm{gl}}\, I_1(t) + c_{\mathrm{gl}}\, I_0(t) + H_0^{\mathrm{gl}},
	\label{eq:glacier_level}
\end{equation}
where $I_2(t) = \int_{t_0}^{t} T(\tau)^2\, d\tau$, $I_1(t) = \int_{t_0}^{t} T(\tau)\, d\tau$, and $I_0(t) = t - t_0$ are the design vectors precomputed from the monthly temperature record, and $H_0^{\mathrm{gl}}$ is the initial sea level offset at the baseline epoch.

\textbf{Priors:} We assign the following priors to the glacier parameters:
\begin{align}
	a_{\mathrm{gl}} &\sim \mathcal{N}\!\left(0,\ 1.82\ \mathrm{mm\,yr^{-1}\,{^\circ C}^{-2}}\right), \label{eq:glacier_prior_a} \\
	b_{\mathrm{gl}} &\sim \mathrm{HalfNormal}\!\left(\sigma = 5\ \mathrm{mm\,yr^{-1}\,{^\circ C}^{-1}}\right), \label{eq:glacier_prior_b} \\
	c_{\mathrm{gl}} &\sim \mathcal{N}\!\left(0.3,\ 2.0\ \mathrm{mm\,yr^{-1}}\right), \label{eq:glacier_prior_c} \\
	\sigma_{\mathrm{extra}}^{\mathrm{gl}} &\sim \mathrm{HalfNormal}\!\left(\sigma = 2\ \mathrm{mm}\right), \label{eq:glacier_prior_sigma} \\
	H_0^{\mathrm{gl}} &\sim \mathcal{N}\!\left(0,\ 10\ \mathrm{mm}\right). \label{eq:glacier_prior_H0}
\end{align}
Unlike the thermosteric and Greenland components, the prior on $a_{\mathrm{gl}}$ is symmetric about zero rather than restricted to positive values. Glaciers are a finite, depleting reservoir: under sustained warming, lower-elevation glaciers disappear entirely while high-elevation accumulation zones receive increased precipitation, causing the aggregate mass-loss rate to peak and then decline. This reservoir depletion and the associated hypsometric feedback (glacier retreat to higher, colder elevations) would manifest as $a_{\mathrm{gl}} < 0$, which a one-sided prior would exclude by construction. The scale is set so that $P(|a_{\mathrm{gl}}| > 3\ \mathrm{mm\,yr^{-1}\,{^\circ C}^{-2}}) = 0.10$, providing weak regularization toward the linear model while allowing the data to determine the sign and magnitude of any nonlinearity. The half-normal prior on $b_{\mathrm{gl}}$ enforces positive sensitivity: glaciers lose mass in a warming climate. The prior on $c_{\mathrm{gl}}$ is centered near zero with sufficient width to accommodate a positive background trend from post--Little Ice Age disequilibrium.

\textbf{Data and calibration:} The model is calibrated against the GlaMBIE global glacier consensus estimate \citep[2000--2023;][]{GlaMBIE2024}, which combines glaciological field measurements, satellite altimetry, DEM differencing, and GRACE/GRACE-FO gravimetry across all 19 glacier regions of the Randolph Glacier Inventory. We use GlaMBIE rather than the glacier component of the \citet{Frederikse2020} budget reconstruction because the latter relies on extrapolating a small sample of directly measured glaciers to the global total via area--volume scaling for the pre-satellite era (1961--2000), introducing model dependence that we seek to avoid. The GlaMBIE record is shorter (24 years versus 57) but is entirely observational: every annual estimate is derived from direct measurements with near-global coverage. This is a critical distinction from the Greenland SMB data (Section~\ref{sec:greenland}), where the observational record is itself the output of a regional climate model (RACMO); for glaciers, the rate--temperature fit extracts an empirical sensitivity directly from observations without a process-model intermediary. The observation model is
\begin{equation}
	H_{\mathrm{gl}}^{\mathrm{obs}}(t_i) \sim \mathcal{N}\!\left(H_{\mathrm{gl}}(t_i),\ \sigma_{\mathrm{obs}}(t_i)^2 + \left(\sigma_{\mathrm{extra}}^{\mathrm{gl}}\right)^2\right),
	\label{eq:glacier_likelihood}
\end{equation}
where $\sigma_{\mathrm{obs}}(t_i)$ is the time-varying observational uncertainty from the GlaMBIE consensus and $\sigma_{\mathrm{extra}}^{\mathrm{gl}}$ is a learned model-inadequacy term that absorbs any systematic misfit between the rate--temperature model and the observations not accounted for by measurement uncertainty (e.g., missing covariates, wrong functional form, temporally correlated residuals). This parameter measures the detectable component of structural uncertainty within the calibration window; it does not bound the extrapolation error at temperatures outside the calibration range, which is a distinct and generally larger source of uncertainty in conditional predictions. Posterior samples are drawn using an affine-invariant ensemble MCMC sampler.

\textbf{Volume cap:} Glaciers are a finite ice reservoir with a total volume of approximately $0.32$~m SLE \citep{Farinotti2019}. When projecting the fitted rate--temperature relationship forward under warming scenarios, we impose a hard ceiling: the cumulative glacier contribution to sea level cannot exceed the total available glacier volume. This is applied per Monte Carlo sample, so the conditional predictive distribution at each future time step reflects the physical constraint that mass loss must cease when no ice remains. This cap is distinct from the hypsometric feedback (glacier retreat to higher elevations reducing the effective temperature sensitivity), which has been documented at the individual glacier scale \citep{Marzeion2014} but has not been demonstrated as a globally significant effect over the observational period. The symmetric prior on $a_{\mathrm{gl}}$ allows the data to detect a declining sensitivity ($a_{\mathrm{gl}} < 0$) if the hypsometric feedback is operating at the global aggregate level; the posterior is centered near zero ($a_{\mathrm{gl}} = -0.12\ [-0.60,\ 0.37]$~mm\,yr$^{-1}$\,$^\circ$C$^{-2}$), indicating that the 2000--2024 data do not require nonlinearity in either direction.

\textbf{Validation:} The calibrated glacier model is validated against the glacier component of the \citet{Frederikse2020} budget reconstruction (1900--2018) and the satellite-era glacier component from \citet{Horwath2022} (1993--2016), both rebased to match GlaMBIE at the 2005 baseline to remove systematic offsets from different processing chains and peripheral glacier treatment. The \citet{Frederikse2020} reconstruction extends further back in time than the GlaMBIE calibration window and provides a check on whether the fitted sensitivity extrapolates to pre-2000 conditions. We also compare our posterior sensitivity $b_{\mathrm{gl}}$ against the emergent temperature sensitivity from the PyGEM process-based glacier model \citep{Rounce2023}, which reported an approximately linear relationship between cumulative glacier mass loss and global warming level across $+1.5$ to $+4\,^\circ$C. Our data-driven sensitivity serves as an independent observational benchmark for the process model: agreement would increase confidence in both approaches, while disagreement would indicate where the process model departs from what the observations support.

	\subsubsection{Greenland}
	\label{sec:greenland}

Greenland contributes to sea level rise through surface mass balance (SMB) anomalies driven by atmospheric warming and ice-dynamic discharge driven by ocean thermal forcing at marine-terminating glacier margins. We treat these separately: SMB is taken directly from observations, while discharge is modeled with a physically motivated time-delay model.

\paragraph{Surface mass balance.}
The \citet{Mouginot2019} SMB record (1972--2018) is derived from the RACMO regional climate model \citep{Noel2018}, which computes surface energy balance, snowfall, refreezing, and runoff on a high-resolution grid forced by ERA reanalysis. We use this record directly as observed data rather than fitting a rate--temperature polynomial. Fitting a simpler model to RACMO output would discard information: RACMO already resolves the spatially heterogeneous processes (albedo feedback, rain/snow partitioning, refreezing capacity) that determine SMB, and a polynomial in temperature cannot recover what the regional climate model computes. This is distinct from the glacier component, where GlaMBIE \citep{GlaMBIE2024} provides direct observations without a process-model intermediary, making a data-driven rate--temperature fit the appropriate approach.

For projections, the SMB anomaly relative to the baseline period scales with temperature \citep[][Eq.~4.8]{CuffeyPaterson2010}:
\begin{equation}
	\Delta \dot{H}_{\mathrm{smb}} = C_T\, \Delta T_{\mathrm{global}} + C_{T^2}\, \Delta T_{\mathrm{global}}^2,
	\label{eq:smb_sensitivity}
\end{equation}
where $C_T$ is the linear sensitivity incorporating both increased runoff and the partial offset from Clausius--Clapeyron scaling of accumulation \citep{Gregory2006}, and $C_{T^2}$ captures the nonlinear enhancement from ablation zone expansion under sustained warming \citep{Noel2021, Sellevold2020}. Published RCM sensitivities are reported per degree of local Greenland temperature; we convert to GMST using an Arctic amplification factor of approximately 2.0. Multiple RCMs constrain the linear sensitivity to $C_T \approx -200$ to $-250$\,Gt\,yr$^{-1}$\,$^\circ$C$^{-1}$ of GMST: MAR gives $-100$ to $-150$\,Gt\,yr$^{-1}$\,$^\circ$C$^{-1}$ of local temperature \citep{Fettweis2013, Hanna2021}, RACMO gives approximately $-100$\,Gt\,yr$^{-1}$\,$^\circ$C$^{-1}$ of local temperature \citep{Noel2018}, with HIRHAM producing values similar to MAR. We use the inter-RCM spread as structural uncertainty: under identical SSP5-8.5 forcing, end-of-century SMB ranges from $-964$\,Gt\,yr$^{-1}$ (RACMO) to $-1735$\,Gt\,yr$^{-1}$ (MAR), a factor of $\sim\!1.8$ \citep{Glaude2024}. We draw
\begin{align}
	C_T &\sim \mathcal{N}\!\left(-200,\, 80^2\right)\,\mathrm{Gt\,yr^{-1}\,{^\circ C}^{-1}\ GMST}, \label{eq:smb_CT} \\
	C_{T^2} &\sim \mathcal{N}\!\left(-50,\, 30^2\right)\,\mathrm{Gt\,yr^{-1}\,{^\circ C}^{-2}\ GMST}, \label{eq:smb_CT2}
\end{align}
and propagate this uncertainty through the Monte Carlo ensemble. The quadratic term reflects the approximately sixfold acceleration in SMB decline above $\sim\!2.7\,^\circ$C GMST documented in high-resolution RCM experiments \citep{Sellevold2020, Noel2021}. As an independent check, we compute a data-driven $C_T$ from GRACE total mass balance minus satellite-era discharge \citep{Otosaka2023, Mankoff2021}, which yields $C_T = 177 \pm 65$\,Gt\,yr$^{-1}$\,$^\circ$C$^{-1}$ of GMST ($R^2 = 0.21$, 29 annual points). The low $R^2$ reflects high interannual variability from weather (NAO, atmospheric blocking) rather than a poor trend estimate; the data-driven value is within $1\sigma$ of the RCM-derived central estimate and serves as an observational lower bound.

\paragraph{Discharge model.}
The cumulative Greenland sea level contribution is
\begin{equation}
	H_{\mathrm{GrIS}}(t) = H_{\mathrm{smb}}^{\mathrm{obs}}(t) + H_{\mathrm{dyn}}(t),
	\label{eq:greenland_total}
\end{equation}
where the discharge component is modeled as a demeaned time-delay response to subsurface ocean temperature:
\begin{equation}
	\Delta H_{\mathrm{dyn}}(t) = \gamma \left[\int_{t_0}^{t} T_{\mathrm{ocean}}(t' - \delta)\, dt' - \theta\right] + r_0\, [t - \bar{t}],
	\label{eq:greenland_dyn}
\end{equation}
where $\Delta H_{\mathrm{dyn}}(t) = H_{\mathrm{dyn}}(t) - \langle H_{\mathrm{dyn}} \rangle_{\mathrm{cal}}$ is the demeaned cumulative discharge, $T_{\mathrm{ocean}}(t - \delta)$ is the subsurface ocean temperature (EN4, 200--500\,m, Greenland peripheral waters; \citealp{Good2013}) lagged by $\delta$~years, $\theta = \langle \int_{t_0}^{t} T_{\mathrm{ocean}}(t' - \delta)\, dt' \rangle_{\mathrm{cal}}$ is the calibration-period mean of the cumulative integral, and $\bar{t}$ is the calibration-period mean time. The sensitivity $\gamma$ captures the rate of sea-level rise per degree of delayed ocean thermal forcing, and $r_0$ is a background secular drift. The time delay $\delta$ represents the lag between subsurface ocean warming and the grounding-line/calving-front response, encompassing fjord renewal, submarine melt, and glacier dynamic adjustment \citep{Straneo2015, Felikson2017}. Demeaning eliminates the baseline offset $H_0^{\mathrm{dyn}}$ without loss of information, since projections are always relative to a baseline epoch.

\paragraph{Independence of SMB and discharge.}
The model treats SMB and discharge as independent, additive components. Three lines of evidence support this structure. First, \citet{King2020} showed that Greenland's discharge acceleration (2000--2005) was driven by calving-front retreat, with a consistent 4--5\% speedup per km of retreat that has persisted independent of interannual surface melt variability. Second, dynamic thinning propagates inland from the terminus at rates of $\sim$10\,m\,yr$^{-1}$, an order of magnitude larger than SMB-driven thinning in the ablation zone \citep{Felikson2017}; these are spatially separated processes. Third, a potential SMB--discharge feedback through driving stress ($\tau_d = \rho_{\mathrm{ice}} g h \nabla h$)---in which SMB-driven thinning via the kinematic boundary condition $\partial h/\partial t = \dot{b} - \nabla \cdot (\mathbf{u} h)$ modifies glacier dynamics---is not supported by the data. Using the per-glacier records from \citet{Mouginot2019} for all 219 marine-terminating glaciers, we find no significant correlation between area-normalized cumulative SMB deficit and discharge change ($r = +0.12$, $p = 0.08$; wrong sign for the feedback hypothesis) or between cumulative mass balance and discharge change ($r = -0.01$, $p = 0.93$). The meltwater--sliding feedback is further constrained by observations at land-terminating sectors showing that increased melt can produce \emph{deceleration} through efficient subglacial drainage development \citep{Tedstone2015}. Under sustained future warming, cumulative thinning may eventually become large enough to affect discharge, but this is not detectable in the 1972--2018 record.

\paragraph{Delay selection and calibration.}
The time delay $\delta$ is determined in two stages. First, cross-correlation of the EN4 subsurface ocean temperature rate against the \citet{Mouginot2019} discharge rate identifies a peak at $\delta = 6$\,yr ($r = 0.84$). Second, for each candidate $\delta \in \{4, 5, 6, 7, 8\}$\,yr, we fit $\gamma$ and $r_0$ by weighted least squares (WLS) on the demeaned cumulative observations, with weights $w_j = 1/\sigma_{\mathrm{dyn}}(t_j)^2$. BIC selects $\delta = 5$\,yr ($R^2 = 0.995$). As a sanity check, the observed end-of-record discharge rate (estimated over the last 10~years of the \citet{Mouginot2019} record) is appended to the WLS system as an additional equation at its natural measurement uncertainty; this does not materially affect the fit but verifies that the cumulative model produces a physically reasonable departure rate.

For a linear model with Gaussian likelihood and known observation variances, the WLS solution is the Bayesian posterior under flat priors: the posterior mean is the WLS estimate $(X^\top W X)^{-1} X^\top W y$ and the posterior covariance is $(X^\top W X)^{-1}$. Monte Carlo samples are drawn from this Gaussian posterior, with BIC-weighted mixing across candidate delays to propagate delay uncertainty. Each Monte Carlo sample draws a single realization of the physical parameters $(\gamma, r_0, \delta)$ and the ocean transfer function $(\alpha, \beta)$, then evaluates all SSP scenarios under that draw. This ensures that inter-SSP differences in the discharge ensemble isolate the effect of the emissions pathway rather than conflating it with parameter uncertainty.

The model is calibrated against the discharge component of \citet{Mouginot2019} (1972--2018, 47~annual observations), forced by EN4 subsurface ocean temperature (200--500\,m, 58--80$^\circ$N, 75--5$^\circ$W; \citealp{Straneo2013}). The independent \citet{Mankoff2021} product (1986--2023) and NOAA Arctic Report Card discharge estimates \citep{Moon2024, Moon2025} are withheld for validation. Total Greenland mass balance (observed SMB + modeled discharge) is validated against IMBIE \citep{Otosaka2023} and \citet{Horwath2022}.

	\subsubsection{Antarctica: Peninsula and EAIS}

The Antarctic Peninsula and East Antarctic Ice Sheet (EAIS) are smaller contributors to GMSL than Greenland, glaciers, or WAIS, but are treated explicitly to close the component budget. EAIS projections use literature-derived SMB sensitivities for the same reason as Greenland SMB: the observational record is too short and noisy for a data-driven temperature regression, and the underlying accumulation physics requires spatially resolved atmospheric modeling. The Peninsula is modeled with an empirical linear fit that captures the combined SMB and discharge signal directly.

\paragraph{East Antarctic Ice Sheet.}
EAIS mass balance is dominated by accumulation: warming increases snowfall via Clausius--Clapeyron scaling ($\sim\!5\%\,^\circ$C$^{-1}$ of $\sim\!1200$~Gt\,yr$^{-1}$ total accumulation), partially offset by marginal dynamic losses. We use the sensitivity $C_T^{\mathrm{EAIS}} \sim \mathcal{N}(+60,\, 20^2)$\,Gt\,yr$^{-1}$\,$^\circ$C$^{-1}$ of GMST from \citet{Frieler2015} and \citet{Ligtenberg2013}. Because $C_T > 0$ (mass gain under warming), EAIS produces a \emph{negative} sea level contribution that partially offsets other components. As a diagnostic, we fit a Bayesian rate--temperature model to the IMBIE East Antarctic record (1992--2020; \citealp{Otosaka2023}); BIC model selection confirms the trend-only model over the quadratic, consistent with no detectable temperature sensitivity over the 28-year satellite era.

\paragraph{Antarctic Peninsula.}
The Peninsula contributes $\sim\!45$~mm to GMSL by 2100 under SSP2-4.5, approximately 9\% of the total---small relative to other components but not negligible. We model the Peninsula with a Bayesian linear rate--temperature fit to IMBIE (1992--2020), which captures the combined SMB and discharge signal in a single empirical relationship:
\begin{equation}
	\dot{H}_{\mathrm{Pen}}(t) = b_{\mathrm{Pen}}\, T(t) + c_{\mathrm{Pen}},
	\label{eq:peninsula_rate}
\end{equation}
where $b_{\mathrm{Pen}}$ is the net temperature sensitivity and $c_{\mathrm{Pen}}$ is a background mass-loss rate. BIC model selection confirms the linear model over a quadratic. This approach is justified on two grounds. First, the Peninsula signal is dominated by ice dynamics ($\sim\!94\%$ of mass loss on decadal timescales; \citealp{Kim2024}), so the linear fit is primarily capturing discharge, which is the component that would require a separate ODE in a Greenland-style decomposition. Second, while recent work has documented a widespread discharge acceleration on the western Peninsula since 2018 driven by subsurface ocean warming \citep{Davison2024}, with discharge trends tripling from $\sim\!50$ to $\sim\!160$~Mt\,yr$^{-2}$ \citep{Davison2024}, and satellite observations confirm sustained velocity increases through 2023 \citep{Wuite2025, Wallis2023}, the total contribution remains small enough that the $\sim\!15$~mm difference between the linear extrapolation and a more physically decomposed model is well within the uncertainty of larger components (the WAIS 90\% credible interval alone spans $\sim\!250$~mm). The comprehensive grounding-line discharge dataset of \citet{Rignot2025}, which provides monthly estimates for 998 basins from 1996--2024, notes that Peninsula discharge estimates show less inter-method agreement than for West Antarctica, further limiting the value of a more complex decomposition at this stage.

We note that the post-2018 acceleration, if sustained, would add $\sim\!15$~mm to our central estimate by 2100---a correction we flag for future work but do not incorporate here given the short observational baseline of the acceleration ($\sim\!5$ years) and its small magnitude relative to total GMSL uncertainty.

\subsection{WAIS uncertainty}
\label{sec:wais}

The West Antarctic Ice Sheet dominates sea level projection uncertainty at centennial timescales. Unlike thermosteric, glacier, and Greenland components, WAIS dynamics cannot be modeled with a data-driven rate--temperature regression: the observational record (28 years of IMBIE satellite gravimetry) is too short to constrain a temperature-dependent response for a system whose instabilities unfold over decades to centuries, and the physics of marine ice sheet instability involves threshold processes that are poorly represented by smooth functional forms. We therefore adopt a scenario-based approach that constructs a probability distribution over WAIS contributions from a discrete mixture of physically distinct regimes, each with literature-informed ranges and an explicit skewness parameterization motivated by the stochastic dynamics of grounding-line retreat.

\subsubsection{A4 scenario framework}

We define three scenarios for WAIS contributions at 2100, reflecting distinct physical regimes of ice sheet behavior:

\begin{enumerate}
	\item[\textbf{S1.}] \textbf{Status quo} ($P = 0.10$; 25--85\,mm). MISI does not operate---bed topography, m\'{e}lange buttressing, and ice-shelf re-formation prevent the runaway grounding-line flux feedback from activating, regardless of ocean warming. Discharge responds linearly to ocean thermal forcing without instability amplification. The range is derived from observed discharge sensitivity and projected ocean warming: \citet{Joughin2021} showed that Pine Island Glacier ice loss scales linearly with time-averaged melt volume ($r^2 \geq 0.98$), and that melt depends quadratically on thermal forcing. Under projected Amundsen shelf warming of $+0.7$ to $+1.5\,^\circ$C by 2100 \citep{Naughten2023}, the melt enhancement factor is $f = [(\Delta T_0 + \Delta T_{\mathrm{warm}})/\Delta T_0]^2 \approx 1.7$--$2.9$ (consistent with \citealp{Jourdain2022}: $1.4$--$2.2\times$ under RCP8.5). The lower bound (25\,mm) is the IMBIE 1992--2020 time-averaged rate ($\sim\!0.24$\,mm\,yr$^{-1}$; \citealp{Otosaka2023}) sustained over 95 years. The upper bound (85\,mm) ramps the peak IMBIE rate ($\sim\!0.44$\,mm\,yr$^{-1}$) to $0.44 \times 2.9 \approx 1.3$\,mm\,yr$^{-1}$ by 2100, giving $(0.44 + 1.3)/2 \times 95 \approx 83$\,mm. The low weight (10\%) reflects growing evidence that MISI may already be underway: grounding-line retreat on retrograde beds is observed at Thwaites Glacier \citep{Joughin2014}, ocean thermal forcing sufficient to drive WAIS retreat is committed regardless of emissions pathway \citep{Naughten2023}, and present-day forcing held constant is sufficient to deglaciate large parts of WAIS \citep{vandenAkker2025}.

	\item[\textbf{S2.}] \textbf{MISI} ($P = 0.80$; 150--1000\,mm; $\alpha = +4$). Marine ice sheet instability is triggered, with the full range of amplifying processes that current models omit. The lower bound (150\,mm) represents late-onset MISI ($\sim$2060) with high residual buttressing, limited primarily to the Thwaites basin. Two independent arguments anchor this value: (i)~rate separation from S1---S1's upper bound (85\,mm) is the maximum without MISI, so the instability feedback must produce rates clearly above the melt-driven linear response, and even late-triggering MISI that doubles the late-century rate gives $\sim\!150$\,mm; (ii)~bed geometry---if MISI drives Thwaites' grounding line through even a portion of its retrograde bed section (bed deepening from $\sim\!700$\,m to $\sim\!2000$\,m; BedMachine, \citealp{Morlighem2020}), partial traverse produces $\sim\!50$--60\,mm\,SLE from Thwaites alone, which combined with melt-driven contributions from other basins exceeds 150\,mm. The upper bound (1000\,mm) matches the Bamber et al.\ structured expert judgment 95th percentile for WAIS under $+5\,^\circ$C \citep[$\sim\!930$\,mm;][]{Bamber2019}. The skewness parameter $\alpha = +4$ is constrained by two requirements: (i)~\citet{Robel2019} showed analytically that MISI produces substantial positive skewness when the grounding-line flux nonlinearity exponent exceeds 3, prescribing $\alpha > 0$; (ii)~the \citet{Bamber2019} conditional-on-MISI median ($\sim\!250$--300\,mm, obtained by removing the $\sim\!15\%$ of outcomes corresponding to stability and $\sim\!5\%$ to MICI extremes from the unconditional distribution) must be reproduced at the $n_{\mathrm{ref}} = 3$ reference rheology. With bounds of 150--1000\,mm, $\alpha = 4$ produces a pre-correction median of $\sim\!302$\,mm, consistent with the Bamber conditional range. After rheology correction ($\times\,1.28$), the median shifts to $\sim\!387$\,mm---higher than Bamber, as expected because the correction captures the $n = 3 \to 4$ bias that postdates the 2018--2019 elicitation \citep{Millstein2022, Martin2026}.

	This scenario merges what were previously separate ``moderate MISI'' and ``MISI with amplifiers'' scenarios. The distinction is not physically defensible: if MISI is triggered, the amplifying processes that models omit---calving \citep{Aschwanden2021, Fricker2025}, subglacial hydrology that can amplify discharge up to threefold \citep{Dow2022}, and the 21--35\% underestimation from incorrect ice rheology ($n = 3$ vs.\ observed $n \approx 4$; \citealp{Martin2026, GetraerMorlighem2025})---operate concurrently. There is no physical basis for a scenario in which MISI proceeds exactly as current models predict, because those models do not represent the processes that govern the instability rate.

	\item[\textbf{S3.}] \textbf{MISI + MICI} ($P = 0.10$; 600--1400\,mm; $\alpha = -3$). Marine ice cliff instability \citep{Bassis2012, DeContoP2016} operates in addition to MISI. The lower bound (600\,mm) reflects the Edwards et al.\ emulator estimate for Antarctic SLR with MICI under RCP8.5 \citep[$\sim\!450$\,mm most likely, total Antarctic;][]{Edwards2019}, augmented by MISI contributions. The upper bound (1400\,mm) is anchored to the total ice volume above flotation in the Amundsen Sea Embayment ($V_{\mathrm{af}} \approx 1100$\,mm; BedMachine Antarctica, \citealp{Morlighem2020}), corrected for the $n = 3 \to 4$ rheology bias ($\times\,1.28$; \citealp{Martin2026, GetraerMorlighem2025}), representing near-complete ASE deglaciation---the maximum WAIS contribution achievable within the century without requiring substantial mass loss from non-ASE basins (Siple Coast, Marie Byrd Land) that lack the retrograde bed geometry and sustained ocean thermal forcing to support rapid retreat. This value is consistent with the \citet{Bamber2019} WAIS 95th percentile (930\,mm) after rheology correction ($\sim\!1190$\,mm) and exceeds the most aggressive published model with MICI \citep[$\sim\!640$--$1140$\,mm total Antarctic under RCP8.5;][]{DeContoP2016}. We assign $\alpha = -3$ (negative skewness in log-space, concentrating probability toward the lower portion of the range), reflecting the accumulating evidence against MICI viability during the 21st century: \citet{Morlighem2024} showed in a multi-model study that Thwaites would not retreat further due to MICI and that calving rates would need to be $\geq\!25\times$ higher than physically motivated models suggest; \citet{Clerc2019} demonstrated that realistic ice-shelf removal timescales raise the critical cliff height from $\sim\!90$\,m to $\sim\!540$\,m; and \citet{Schlemm2022} found MICI to be self-limiting due to m\'{e}lange buttressing. IPCC AR6 assigns low confidence to MICI \citep{Fox-Kemper2021}.
\end{enumerate}

\noindent Table~\ref{tab:wais_scenarios} summarizes the scenario parameters and the resulting mixture distribution.

\begin{table}[ht]
\centering
\caption{WAIS A4 scenario framework. Ranges denote the 5th--95th percentile bounds used to set the skew-normal location and scale in log-space. The skewness parameter $\alpha$ controls the asymmetry of the within-scenario distribution ($\alpha > 0$: heavy right tail; $\alpha < 0$: heavy left tail; $\alpha = 0$: log-normal). The trajectory exponent $\beta$ governs the temporal structure of mass loss: $H(t) = H_{2100}\cdot[(t - t_0)/95]^\beta$, where $\beta = 1$ is linear (constant rate) and $\beta > 1$ produces accelerating (back-loaded) trajectories consistent with the grounding-line flux scaling $q_g \propto h_g^{n+1}$ on retrograde beds \citep{Schoof2007}. A multiplicative rheology correction ($\times\,1.28 \pm 0.07$) is applied to all scenarios. The combined A4 distribution is SSP-independent.}
\label{tab:wais_scenarios}
\smallskip
\begin{tabular}{l c c c c c l}
\toprule
Scenario & Weight & Range (mm) & $\alpha$ & $\beta$ & Distribution & Physical regime \\
\midrule
S1: Status quo   & 0.10 & 25--85     & 0    & 1.0 (fixed) & Log-normal  & No MISI; linear melt-driven \\
S2: MISI         & 0.80 & 150--1000  & $+4$ & LN(1.8, 0.3) & Right-skewed & MISI; accelerating \\
S3: MISI + MICI  & 0.10 & 600--1400  & $-3$ & LN(2.2, 0.3) & Left-skewed  & MISI + MICI; strongly accel. \\
\midrule
\multicolumn{7}{l}{\textit{Post-hoc correction applied to all scenarios:}} \\
Rheology ($n=4$) & ---  & $\times\,1.28\ [1.14,\,1.42]$ & --- & --- & Gaussian & Glen's law $n=3 \to 4$ bias \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Distribution and corrections}

Within each scenario, samples are drawn from a skew-normal distribution in log-space. This parameterization has three transparent parameters: a location $\xi$ and scale $\omega$ (determined by the requirement that the 5th and 95th percentiles match the scenario bounds) and a shape parameter $\alpha$ that controls skewness. Setting $\alpha = 0$ recovers a log-normal; $\alpha > 0$ produces positive skew (heavy right tail); $\alpha < 0$ produces negative skew. The skew-normal family is chosen because $\alpha$ maps directly to the physical insight of \citet{Robel2019}: MISI creates a systematic asymmetry in which the probability of ice loss exceeding the most likely outcome is substantially greater than the probability of falling below it.

\paragraph{Trajectory exponent.}
The temporal distribution of mass loss within the century is governed by a power-law ramp $H(t) = H_{2100} \cdot [(t - t_0)/(2100 - t_0)]^\beta$, where $\beta$ is drawn per Monte Carlo sample from a scenario-dependent log-normal distribution (Table~\ref{tab:wais_scenarios}). This parameterization derives from the grounding-line flux condition $q_g \propto h_g^{n+1}$ \citep{Schoof2007}: on a retrograde bed, retreat increases the ice thickness at the grounding line, which increases flux, which drives further retreat. Dimensional analysis of this coupled system shows that cumulative mass loss grows as $(t - t_0)^\beta$ with $\beta > 1$ during the acceleration phase, where the exponent depends on the flux nonlinearity ($n$) and bed slope \citep{Schoof2012, Robel2019}.

For S1 (no MISI), $\beta = 1$ (constant rate). For S2 (MISI), $\beta$ has a median of 1.8: the 2050 cumulative is $\sim\!30\%$ of the 2100 value rather than the $47\%$ that a linear ramp would produce. This back-loading is consistent with the Robel et al.\ ensemble trajectories and with the \citet{Bamber2019} structured expert judgment, in which experts placed only $\sim\!22\%$ of WAIS loss before 2050. The spread in $\beta$ (90\% CI approximately 1.1--3.0) reflects uncertainty in bed geometry across WAIS drainage basins: Thwaites has a deep retrograde bed that supports large $\beta$, while other basins may have flatter or prograde beds that limit acceleration.

The trajectory exponent and the endpoint skewness $\alpha$ encode complementary aspects of the same flux-geometry physics: $\alpha$ captures how the grounding-line nonlinearity amplifies cross-ensemble uncertainty toward worse outcomes, while $\beta$ captures how it concentrates within-trajectory mass loss toward the end of the century. A framework that includes $\alpha$ but assumes $\beta = 1$ would be internally inconsistent---acknowledging the flux nonlinearity for uncertainty quantification but ignoring it for temporal structure.

\paragraph{Rheology correction.}
A multiplicative rheology correction is applied to all samples from all scenarios. Current ice sheet models universally assume Glen's flow law exponent $n = 3$, but satellite observations yield $n = 4.1 \pm 0.4$ \citep{Millstein2022}, supported by field measurements \citep{Bons2018}, laboratory meta-analysis \citep{Fan2025}, and a theoretical framework connecting laboratory to ice-sheet scales \citep{Ranganathan2024}. \citet{Martin2026} showed that this mismatch causes a 21--35\% underestimation of 100-year SLR in the MISMIP+ benchmark, while \citet{GetraerMorlighem2025} found $32 \pm 14\%$ greater Amundsen Sea Embayment loss with $n = 4$ in the ISMIP6 framework. We apply the correction as a multiplicative factor drawn from $\mathcal{N}(1.28,\, 0.07^2)$, truncated below at 1.0, where the median spans the range from the MISMIP+ result (1.21) to the ABUMIP result (1.35).

The combined A4 distribution is SSP-independent: WAIS dynamics on projection timescales are driven primarily by ocean thermal forcing with long lag times, not by contemporaneous atmospheric warming. \citet{Naughten2023} showed that rapid West Antarctic ice-shelf melting at $\sim\!3\times$ historical rates is unavoidable regardless of emissions scenario.

\subsubsection{Independent validation: Bamber et al.\ SEJ}

As an independent cross-check, we compare the A4 mixture against the structured expert judgment (SEJ) of \citet{Bamber2019}, which elicited WAIS-specific probability distributions from 22 ice sheet experts using Cooke's classical model. Under $+5\,^\circ$C, their WAIS 5th--95th percentile range is $-50$ to $930$\,mm (median 180\,mm). Our A4 mixture produces a median in the hundreds of mm with a strongly right-skewed distribution---wider than the Bamber distribution, as expected given that the SEJ predates the quantification of the rheology bias \citep{Martin2026} and recent observational evidence of ongoing destabilization. The broad agreement in distributional shape between two methodologically independent approaches---physics-informed scenario mixture vs.\ performance-weighted expert judgment---strengthens confidence that the ISMIP6 medium-confidence range substantially underestimates WAIS uncertainty.

\section{Results}

\subsection{Semi-empirical forecasts}

\subsection{Physics-informed forecasts}
	\subsubsection{Thermosteric}

\paragraph{Hindcast rates.}
The full-depth observational hindcast produces thermosteric rates consistent with the multi-product literature: $\sim\!1.2$\,mm\,yr$^{-1}$ over 1993--2020 (WCRP ensemble: $1.1$--$1.3$\,mm\,yr$^{-1}$; \citealp{vonSchuckmann2023}) and $\sim\!0.7$\,mm\,yr$^{-1}$ over 1955--2010. The 0--2000\,m NOAA product (2005--2025) confirms a visible divergence from the 0--700\,m record in the Argo era, consistent with the 700--2000\,m layer contributing $\sim\!25$--$30\%$ of total thermosteric rise during this period.

\paragraph{Splice validation.}
The IPCC AR6 projections are level-matched to observations at 2020. In the 2020--2025 overlap window, the shifted IPCC median and the NOAA observations agree within uncertainties, confirming that the splice does not introduce a discontinuity. The IPCC projections implicitly include the full ocean depth through the two-layer energy balance model and CMIP6-derived expansion coefficients, so no additional depth correction is needed for the projection period.

\paragraph{Budget closure.}
A variance-weighted budget attribution diagnostic distributes the residual between the component sum and NASA altimetry (0.37\,mm\,yr$^{-1}$ over 1993--2020) across components in proportion to their uncertainties. Thermosteric absorbs the largest share (57\%), consistent with the known low bias in the NOAA product \citep{Li2022}. The implied shift at 2020 ($+3.2$\,mm, $1\sigma = 2.5$\,mm including the 14\% structural uncertainty) is a $\sim\!1.3\sigma$ adjustment---within the stated uncertainty. All non-ocean components except TWS are within $1.5\sigma$ of their independent estimates, indicating that the component-level framework closes the satellite-era budget without requiring any component to depart from its observational constraints.

	\subsubsection{Glaciers}

The glacier model is calibrated on 24 annual GlaMBIE observations (2000--2024) spanning approximately $1\,^\circ$C of GMST warming relative to the 1995--2005 baseline. The posterior distributions for the rate--temperature parameters are:
\begin{align}
	a_{\mathrm{gl}} &= -0.12\ [-0.60,\ 0.37]\ \mathrm{mm\,yr^{-1}\,{^\circ C}^{-2}}, \label{eq:glacier_result_a} \\
	b_{\mathrm{gl}} &= \phantom{-}0.77\ [0.49,\ 1.05]\ \mathrm{mm\,yr^{-1}\,{^\circ C}^{-1}}, \label{eq:glacier_result_b} \\
	c_{\mathrm{gl}} &= \phantom{-}0.59\ [0.51,\ 0.67]\ \mathrm{mm\,yr^{-1}}, \label{eq:glacier_result_c}
\end{align}
where brackets denote the 90\% credible interval. The model-inadequacy term is small ($\sigma_{\mathrm{extra}}^{\mathrm{gl}} = 0.11\ [0.01,\ 0.34]$~mm), indicating that the rate--temperature model captures the GlaMBIE observations within their stated uncertainties over the calibration period. The fit achieves $R^2 = 0.998$. We emphasize that a small $\sigma_{\mathrm{extra}}$ does not imply small structural uncertainty in conditional predictions at higher temperatures: the model is calibrated over $\sim\!1\,^\circ$C of warming, and extrapolation to $+3$--$4\,^\circ$C carries additional risk from physical processes (reservoir depletion, hypsometric feedback) that may not be detectable within the calibration range. The volume cap and the symmetric prior on $a_{\mathrm{gl}}$ provide partial safeguards, but the out-of-sample structural uncertainty remains the dominant source of predictive uncertainty for the glacier component at centennial horizons.

The posterior on $a_{\mathrm{gl}}$ is centered near zero and spans both positive and negative values (posterior probability $P(a_{\mathrm{gl}} > 0.5) = 1.8\%$), confirming that the data do not require nonlinearity in either direction over the observational period. The linear model ($a_{\mathrm{gl}} = 0$) is sufficient: the instantaneous rate of glacier mass loss is well described by
\begin{equation}
	\dot{H}_{\mathrm{gl}} \approx 0.77\, T + 0.59 \quad \mathrm{mm\,yr^{-1}},
	\label{eq:glacier_linear_rate}
\end{equation}
where $T$ is the GMST anomaly in $^\circ$C relative to the 1995--2005 mean. At the 2024 GMST anomaly of approximately $+1.3\,^\circ$C, this implies a glacier mass-loss rate of $\sim\!1.6$~mm\,yr$^{-1}$, consistent with the GlaMBIE consensus.

The background trend $c_{\mathrm{gl}} = 0.59$~mm\,yr$^{-1}$ represents glacier mass loss at the baseline temperature. This is physically expected: glaciers were losing mass throughout the 20th century due to post--Little Ice Age disequilibrium and the cumulative effects of prior warming.

\paragraph{Comparison with process models.}
The data-driven linear sensitivity $b_{\mathrm{gl}} = 0.77$~mm\,yr$^{-1}$\,$^\circ$C$^{-1}$ provides an observational benchmark for process-based glacier models. Integrating over 85 years (2015--2100), this sensitivity implies approximately 65~mm of cumulative glacier SLR per degree of sustained warming. \citet{Rounce2023}, using the PyGEM process-based glacier evolution model, reported 90--154~mm cumulative glacier loss by 2100 across $+1.5$ to $+4\,^\circ$C warming levels, implying an effective sensitivity of 35--40~mm per degree. Our data-driven value is roughly $1.5$--$2\times$ the process-model value. This discrepancy could reflect the hypsometric feedback incorporated in process models (glacier retreat to higher elevations reduces future sensitivity) or errors in the process models themselves. Process-based glacier models have been calibrated at the individual glacier scale but have not been validated against independent out-of-sample predictions of global aggregate mass loss under warming.  We note that our conditional predictions are bounded by the finite glacier ice volume ($\sim\!0.32$~m SLE), which limits the practical consequence of this sensitivity difference under high-warming scenarios.

An independent diagnostic decomposes the IPCC AR6 glacier median projections into rate--temperature space by fitting $\dot{H}_{\mathrm{IPCC}} = b_{\mathrm{IPCC}}\, T + c_{\mathrm{IPCC}}$ to the decadal rates implied by the IPCC median trajectories. This yields $b_{\mathrm{IPCC}} \approx 0.13$~mm\,yr$^{-1}$\,$^\circ$C$^{-1}$ and $c_{\mathrm{IPCC}} \approx 1.2$~mm\,yr$^{-1}$, stable across SSPs ($R^2 > 0.85$). The structure is inverted relative to our data-driven fit: the IPCC median attributes most of the glacier mass-loss rate to a temperature-independent committed response, with a temperature sensitivity roughly $6\times$ smaller than what GlaMBIE observations support. This structural difference---high committed rate with low sensitivity versus moderate committed rate with high sensitivity---means that the IPCC projections and our model converge at intermediate warming levels but diverge under both low- and high-warming scenarios, with IPCC projecting more loss under stabilization (due to its large $c$) and less under sustained warming (due to its small $b$).

\paragraph{Validation against independent datasets.}
The GlaMBIE-calibrated model is evaluated against the \citet{Frederikse2020} glacier reconstruction (1900--2018) and the \citet{Horwath2022} satellite-era budget (1993--2016), both rebased to match GlaMBIE at the 2005 baseline.  Over the satellite era (2000--2018), all three datasets agree within their stated uncertainties, confirming that the fitted rate--temperature relationship is consistent across independent observational products. Before 2000, the model extrapolation tracks the Frederikse reconstruction, though Frederikse pre-2000 glacier estimates depend on area--volume scaling extrapolations and are therefore not purely observational.

	\subsubsection{Greenland}

The discharge delay model, calibrated against \citet{Mouginot2019} (1972--2018), yields $\gamma = 0.365$\,mm\,yr$^{-1}$\,$^\circ$C$^{-1}$ (90\% CI: 0.352--0.379), $r_0 = 0.104$\,mm\,yr$^{-1}$ (90\% CI: 0.099--0.108), and $\delta = 5$\,yr (BIC-selected). The fitted $R^2 = 0.995$ on the demeaned cumulative discharge. All three parameters agree with independent physical estimates derived from glacier dynamics (Supplementary Material): $\gamma$ and $\delta$ match to within their uncertainties, while $r_0$ exceeds the physical estimate by a factor of $\sim\!1.4$, likely reflecting secular processes not fully captured by the two-source physical derivation.

\paragraph{Validation against withheld data.}
The \citet{Mankoff2021} discharge record (1986--2023) and NOAA Arctic Report Card estimates for 2020 and 2023 \citep{Moon2024, Moon2025} were withheld from calibration. Both fall within the model's 90\% credible interval over their respective periods, confirming that the delay model generalizes beyond the Mouginot calibration window.

\paragraph{Total Greenland mass balance.}
The total hindcast $H_{\mathrm{GrIS}} = H_{\mathrm{smb}}^{\mathrm{obs}} + H_{\mathrm{dyn}}$ is validated against IMBIE \citep{Otosaka2023} and \citet{Horwath2022} over the satellite era. The combined model reproduces the observed acceleration in mass loss from the mid-1990s onward, with the transition from near-balance to sustained loss driven primarily by the discharge response to subsurface ocean warming. SMB remains the dominant term in absolute magnitude ($\sim\!350$\,Gt\,yr$^{-1}$), but discharge variability drives the decadal-scale changes that control the sea-level contribution.

\paragraph{Sensitivity to RCM choice.}
Our hindcast SMB is taken from \citet{Mouginot2019}, which uses RACMO as its sole regional climate model. Comparing against the \citet{Mankoff2021} product, which averages SMB from three RCMs (HIRHAM/HARMONIE, MAR, and RACMO), shows agreement within observational uncertainties over the full 1972--2018 overlap. The three-RCM average is systematically slightly lower, consistent with the $\sim\!15$\,Gt\,yr$^{-1}$ ($\sim\!4\%$) MAR--RACMO offset documented in the GrSMBMIP intercomparison \citep{Fettweis2020}. This difference ($\sim\!0.04$\,mm\,yr$^{-1}$ SLE) is small relative to our other structural uncertainties and does not materially affect the hindcast (Supplementary Material, \S\ref*{sec:rcm_sensitivity}).

	\subsubsection{Antarctica: Peninsula and EAIS}

	\subsubsection{WAIS}

\section{Discussion}

\subsection{Implication of results for sea-level prediction}

\subsection{Forecasting skill} 

\subsection{Calibration considerations}

How far outside the calibration bounds does our model still have skill?

\paragraph{Model inadequacy vs.\ structural uncertainty.}
Each component model includes a learned parameter $\sigma_{\mathrm{extra}}$ that quantifies the systematic misfit between the model and the calibration data beyond what measurement uncertainty explains. A small $\sigma_{\mathrm{extra}}$ indicates that the chosen functional form is adequate for the calibration period and temperature range --- it does not guarantee adequacy outside that range. The distinction matters for interpretation: $\sigma_{\mathrm{extra}}$ is a lower bound on structural uncertainty within sample, not an estimate of the prediction error at temperatures the model has not seen.

For the glacier component, $\sigma_{\mathrm{extra}} \approx 0.1$~mm over 2000--2024, during which GMST warmed by $\sim\!1\,^\circ$C. Extrapolating the fitted linear sensitivity to $+3$--$4\,^\circ$C introduces structural risks --- reservoir depletion, hypsometric feedback, threshold behavior --- that are invisible to $\sigma_{\mathrm{extra}}$ because they have not yet been expressed in the data. The same caveat applies to all components: the Greenland discharge model has been observed for only $\sim\!50$ years of a response time $\tau \sim 10$--$20$~yr. These limitations are inherent to any data-driven framework and are not unique to ours; the advantage of our approach is that the calibration range and the extrapolation lever arm are explicit, whereas process models embed analogous extrapolation assumptions in their parameterizations without reporting them as such.

\paragraph{Budget closure as a diagnostic.}
Each component is calibrated independently against its own observational dataset, and the total GMSL is a diagnostic sum rather than a calibration target. Over the satellite era (1993--2020), the component sum underpredicts NASA altimetry by 0.37\,mm\,yr$^{-1}$, or approximately 12\% of the observed rate. We attribute this residual across components in proportion to their variances: the thermosteric component absorbs the largest share (57\%), consistent with the known low bias in the NOAA product relative to other observational analyses \citep{Li2022, Cheng2017}. TWS absorbs 23\%, reflecting poorly characterized structural uncertainty in groundwater and reservoir reconstructions. All remaining components (glaciers, Greenland, EAIS, Peninsula, WAIS) are within $1.5\sigma$ of their independent estimates, indicating that the budget does not require any of these components to depart from what the observations support. The fact that the residual is attributable to known observational limitations, rather than to missing physics or unmodeled components, provides confidence that the component framework captures the dominant processes governing sea level change.

\subsection{Extensions and next steps}

This work is intended to develop a simple hierarchical, data-driven approach to sea-level rise prediction as a complement to and diagnostic for the complex process-based models that have largely dominated discussions of sea-level rise. 


\bibliographystyle{agufull08}
\bibliography{references}

\end{document}



