\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{enumitem}

\title{Component-Level Sea Level Rise Forecast Models:\\A Brief Literature Assessment}
\author{Literature search compiled 2026-05-10}
\date{}

\begin{document}
\maketitle

\noindent This report identifies the six most relevant published frameworks
for component-level global mean sea level (GMSL) projections---i.e., studies
that decompose total GMSL rise into individual physical contributors
(thermosteric expansion, glaciers, Greenland ice sheet, Antarctic ice sheet,
terrestrial water storage) and sum them probabilistically. Relevance is
assessed relative to a Bayesian, observation-calibrated, component-level
forecasting framework.

%% ========================================================================
\section*{1. Mengel et al.\ (2016)}

\noindent\textbf{Citation:} Mengel, M., Levermann, A., Frieler, K.,
Robinson, A., Marzeion, B., \& Winkelmann, R.\ (2016). Future sea level
rise constrained by observations and long-term commitment.
\textit{Proc.\ Natl.\ Acad.\ Sci.}, 113(10), 2597--2602.\\
\textbf{DOI:} \texttt{10.1073/pnas.1500515113}

\noindent\textbf{Approach.} Each component (thermal expansion, glaciers,
Greenland, Antarctica) is represented by a simple physically motivated
response function with a small number of parameters, independently
calibrated to 20th-century observations via Bayesian inference. Projections
are generated by sampling posteriors and forcing with RCP temperature
scenarios, then summing components. Inter-component correlations arise only
through shared temperature forcing.

\noindent\textbf{Relevance: Very high.} This is the closest published
methodological parallel: per-component semi-empirical models, each
observation-calibrated, summed probabilistically. The main differences are
(i) our use of rate-space and level-space Bayesian models rather than
equilibrium response functions, (ii) our explicit deep-uncertainty
treatment of WAIS, and (iii) our use of a blending step that anchors
projections to satellite-era observed rates.

%% ========================================================================
\section*{2. Kopp et al.\ (2014)}

\noindent\textbf{Citation:} Kopp, R.\,E., Horton, R.\,M., Little, C.\,M.,
Mitrovica, J.\,X., Oppenheimer, M., Rasmussen, D.\,J., Strauss, B.\,H.,
\& Tebaldi, C.\ (2014). Probabilistic 21st and 22nd century sea-level
projections at a global network of tide-gauge sites.
\textit{Earth's Future}, 2(8), 383--406.\\
\textbf{DOI:} \texttt{10.1002/2014EF000239}

\noindent\textbf{Approach.} The foundational probabilistic
component-summation framework. Individual components (thermal expansion,
glaciers, GrIS, AIS, land water storage) are each represented as
probability distributions drawn from IPCC AR5 expert assessment and process
models. Monte Carlo sampling sums components at each location, applying
gravitational fingerprints for regional projections.

\noindent\textbf{Relevance: High.} Established the ``plug-and-play''
component-summation architecture that all subsequent probabilistic
frameworks build on, including IPCC AR6. Our work shares this architecture
but replaces assessment-derived component PDFs with observation-calibrated
Bayesian posteriors, and does not address regional patterns.

%% ========================================================================
\section*{3. Wong et al.\ (2017) --- BRICK}

\noindent\textbf{Citation:} Wong, T.\,E., Bakker, A.\,M.\,R.,
Ruckert, K., Applegate, P., Slangen, A.\,B.\,A., \& Keller, K.\ (2017).
BRICK v0.2, a simple, accessible, and transparent model framework for
climate and regional sea-level projections.
\textit{Geosci.\ Model Dev.}, 10, 2741--2760.\\
\textbf{DOI:} \texttt{10.5194/gmd-10-2741-2017}

\noindent\textbf{Approach.} Couples simple, physically motivated models for
each component (DOECLIM for thermal expansion, Marzeion-type scaling for
glaciers, SMB/dynamics models for GrIS and AIS) within a shared climate
state. The full system is jointly calibrated to observations using MCMC,
producing posteriors that naturally capture inter-component correlations.

\noindent\textbf{Relevance: High.} BRICK is the only published framework
that performs joint Bayesian calibration of all components simultaneously,
making it the closest analogue to our approach in terms of statistical
methodology. A key distinction is that our framework calibrates each
component independently and handles WAIS through a structured scenario
mixture rather than a process model.

%% ========================================================================
\section*{4. Kopp et al.\ (2023) --- FACTS}

\noindent\textbf{Citation:} Kopp, R.\,E., Garner, G.\,G.,
Hermans, T.\,H.\,J., et al.\ (2023). The Framework for Assessing Changes
To Sea-level (FACTS) v1.0: a platform for characterizing parametric and
structural uncertainty in future global, relative, and extreme sea-level
change. \textit{Geosci.\ Model Dev.}, 16, 7461--7489.\\
\textbf{DOI:} \texttt{10.5194/gmd-16-7461-2023}

\noindent\textbf{Approach.} The modular software platform underlying IPCC
AR6 sea level projections. Each component is a pluggable module that can use
different process models or emulators. The framework propagates both
parametric uncertainty (within a model) and structural uncertainty (across
alternative models for the same component).

\noindent\textbf{Relevance: High.} FACTS is the institutional benchmark
against which any new component-level framework will be compared. Our work
can be understood as a complementary approach that replaces the
process-model modules in FACTS with data-driven Bayesian models, trading
physical detail for tighter observational constraint and transparency about
what the data alone can support.

%% ========================================================================
\section*{5. Fox-Kemper et al.\ (2021) --- IPCC AR6 Chapter 9}

\noindent\textbf{Citation:} Fox-Kemper, B., Hewitt, H.\,T., Xiao, C.,
et al.\ (2021). Ocean, Cryosphere and Sea Level Change. In
\textit{Climate Change 2021: The Physical Science Basis} (pp.\ 1211--1362).
Cambridge University Press.\\
\textbf{DOI:} \texttt{10.1017/9781009157896.011}

\noindent\textbf{Approach.} The authoritative component-level assessment,
summing thermal expansion (CMIP6), glaciers (GlacierMIP), Greenland
(ISMIP6 + emulators), Antarctica (ISMIP6 + emulators + structured expert
judgment), and land water storage. Provides both medium-confidence and
low-confidence projections across all SSP scenarios.

\noindent\textbf{Relevance: High.} This is the primary comparison target
for any new SLR projection. Our framework produces systematically higher
medians and wider uncertainty ranges than IPCC AR6 medium confidence,
largely because the IPCC ice sheet component relies on ISMIP6 process
models that may understate dynamic mass loss. The IPCC low-confidence
projections, which incorporate expert judgment, are more consistent with our
results.

%% ========================================================================
\section*{6. Nauels et al.\ (2017) --- MAGICC Sea Level Model}

\noindent\textbf{Citation:} Nauels, A., Meinshausen, M., Mengel, M.,
Lorbacher, K., \& Wigley, T.\,M.\,L.\ (2017). Synthesizing long-term sea
level rise projections --- the MAGICC sea level model v2.0.
\textit{Geosci.\ Model Dev.}, 10, 2495--2524.\\
\textbf{DOI:} \texttt{10.5194/gmd-10-2495-2017}

\noindent\textbf{Approach.} Emulates process-based responses for seven
sub-components (thermal expansion, glaciers, GrIS SMB, GrIS dynamics, WAIS,
EAIS, land water storage) using reduced-complexity parameterizations
calibrated to CMIP5/AR5 output. Embedded within the MAGICC climate model,
ensuring a self-consistent climate trajectory across components. Extends
projections to 2300.

\noindent\textbf{Relevance: Moderate--high.} Shares the component
granularity of our framework (seven sub-components) and the goal of
efficient probabilistic ensemble generation. However, MAGICC calibrates to
process-model output rather than directly to observations, and does not
incorporate deep uncertainty in ice sheet dynamics. Our framework's direct
observational calibration and explicit WAIS scenario structure address both
of these limitations.

%% ========================================================================
\section*{Summary}

\begin{table}[h]
\centering\small
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Study} & \textbf{Calibration source} & \textbf{Components} & \textbf{Relevance} \\
\midrule
Mengel et al.\ (2016)    & Observations (indep.)   & 4 & Very high \\
Kopp et al.\ (2014)      & IPCC assessment         & 5 & High \\
Wong et al.\ (2017)      & Observations (joint)    & 5 & High \\
Kopp et al.\ (2023)      & Process models          & 5+ & High \\
Fox-Kemper et al.\ (2021)& CMIP6 + ISMIP6          & 5 & High \\
Nauels et al.\ (2017)    & CMIP5 emulation         & 7 & Moderate--high \\
\bottomrule
\end{tabular}
\end{table}

\noindent No published framework combines all of the following: (i)
direct Bayesian calibration of each component to observational data,
(ii) a structured deep-uncertainty treatment for ice sheet instabilities,
and (iii) rate-space blending with satellite-era observations to anchor
near-term projections. The closest precedent is Mengel et al.\ (2016),
which shares (i) but not (ii) or (iii).

\end{document}
