\documentclass[11pt]{article}

% Packages
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{cleveref}

% doi command
\newcommand{\doi}[1]{\href{https://doi.org/#1}{doi:#1}}

% Theorem environments
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}

% Custom commands
\newcommand{\dd}{\mathrm{d}}
\newcommand{\Qg}{Q_g}
\newcommand{\qg}{q_g}
\newcommand{\hg}{h_g}
\newcommand{\hhat}{\hat{h}}
\newcommand{\Mhat}{\hat{M}}
\newcommand{\That}{\hat{T}}
\newcommand{\xg}{x_g}
\newcommand{\rhoi}{\rho_{\mathrm{i}}}
\newcommand{\rhow}{\rho_{\mathrm{w}}}
\newcommand{\Aoc}{A_{\mathrm{oc}}}
\newcommand{\Vaf}{V_{\mathrm{af}}}

\title{Bounds on 21st-century WAIS mass loss\\from grounding-line flux theory}
\author{Technical note}
\date{}

\begin{document}
\maketitle

%======================================================================
\section{Introduction}
%======================================================================

The West Antarctic Ice Sheet (WAIS) is the largest source of uncertainty in sea-level projections at centennial timescales. Marine ice sheet instability (MISI) on retrograde beds can drive self-sustaining grounding-line retreat, but the rate and magnitude of this retreat are constrained by ice dynamics and bed geometry. Here we derive two complementary bounds on the WAIS contribution to sea-level rise (SLR) during the 21st century, using the grounding-line flux laws of \citet{Schoof2007}, \citet{Pegler2018}, and \citet{HaseloffSergienko2018}.

\begin{enumerate}
	\item A \textbf{geometric lower bound} ($\sim$200\,mm\,SLE for WAIS): the minimum SLR from grounding-line retreat through the retrograde bed sections, independent of ice rheology ($n$), basal sliding ($m$), and the rate factor ($A$). If MISI is triggered, the bed geometry alone guarantees at least this much mass loss, regardless of the uncertain details of ice dynamics.

	\item A \textbf{buttressing-dependent upper bound}: the maximum SLR achievable under a given level of residual buttressing. In the zero-buttressing limit, the dynamics are so fast ($s > 4.5$, finite-time singularity within decades) that the upper bound saturates at the total ice volume above flotation ($\Vaf \sim 3.3$\,m for WAIS). A non-trivial upper bound requires partial buttressing, parameterized by the framework-specific buttressing exponents of \citet{Pegler2018} and \citet{HaseloffSergienko2018}.
\end{enumerate}

The derivation connects the reduced dynamics of grounding-line flux theory to the scenario ranges of the A4 framework, providing physical grounding for both the lower and upper ends of the MISI scenario (S2: 120--1000\,mm).


%======================================================================
\section{Grounding-line flux law}
\label{sec:flux_law}
%======================================================================

\subsection{The Schoof flux condition}

\citet{Schoof2007} derived the grounding-line flux condition for a marine ice sheet using matched asymptotic expansions. For a power-law sliding law
\begin{equation}
	\tau_b = C\,|u_b|^{m-1}\,u_b,
	\label{eq:sliding_law}
\end{equation}
where $m$ is the velocity exponent ($0 < m \leq 1$; $m \to 0$ for a perfectly plastic bed, $m = 1/n$ for Weertman sliding at Glen's law exponent $n$), the unbuttressed grounding-line flux is
\begin{equation}
	\qg^{\,\mathrm{unbutt}} \propto \hg^{\,s_{\mathrm{full}}},
	\qquad
	s_{\mathrm{full}} = \frac{m + n + 3}{m + 1}.
	\label{eq:s_full}
\end{equation}
Two limiting cases confirm the physical content: for a perfectly plastic bed ($m \to 0$), $s_{\mathrm{full}} \to n + 3$---the flux is highly sensitive to thickness because ice viscosity provides the only resistance; for linear sliding ($m = 1$), $s_{\mathrm{full}} = (n+4)/2$---basal drag provides significant resistance. In both limits $s_{\mathrm{full}} > 2$ for $n \geq 3$.

\subsection{Buttressing and the three frameworks}

Ice-shelf buttressing reduces the flux below its unbuttressed value. Parameterizing the back-stress through a buttressing ratio $\theta \in [0,1]$ ($\theta = 0$: no buttressing; $\theta = 1$: full buttressing), the flux scales as
\begin{equation}
	\qg(\theta) = \qg^{\,\mathrm{unbutt}}\,(1 - \theta)^{s_b},
	\label{eq:buttressed_flux}
\end{equation}
where the \emph{buttressing exponent} $s_b$ depends on the framework:
\begin{align}
	s_b^{\,(S)} &= \frac{n}{m+1} && \text{(Schoof: unconfined)}, \label{eq:sb_schoof} \\
	s_b^{\,(P)} &= \frac{n+1}{m+1} && \text{(Pegler: confined, viscous lateral drag)}, \label{eq:sb_pegler} \\
	s_b^{\,(H)} &= \frac{n+1}{m+1} + \frac{1}{n} && \text{(Haseloff: resolved shear margins)}. \label{eq:sb_haseloff}
\end{align}
For Weertman sliding ($m = 1/n$), \cref{tab:exponents} collects the numerical values. The buttressing exponents $s_b$ control how sensitively the flux responds to partial loss of buttressing; they are distinct from the full flux-thickness exponent $s_{\mathrm{full}}$, which governs the response to grounding-line deepening.

\begin{table}[ht]
\centering
\caption{Flux and buttressing exponents for Weertman sliding ($m = 1/n$). The full flux exponent $s_{\mathrm{full}}$ enters the retreat dynamics; the buttressing exponents $s_b$ enter the buttressing-dependent bound.}
\label{tab:exponents}
\smallskip
\begin{tabular}{l c c c}
\toprule
& $n = 3$, $m = 1/3$ & $n = 4$, $m = 1/4$ & Plastic ($m\to0$) \\
\midrule
$s_{\mathrm{full}} = (m+n+3)/(m+1)$ & 4.75 & 5.80 & $n+3$ \\[4pt]
$s_b^{(S)} = n/(m+1)$ & 2.25 & 3.20 & $n$ \\
$s_b^{(P)} = (n+1)/(m+1)$ & 3.00 & 4.00 & $n+1$ \\
$s_b^{(H)} = (n+1)/(m+1) + 1/n$ & 3.33 & 4.25 & $n+1+1/n$ \\
\bottomrule
\end{tabular}
\end{table}


%======================================================================
\section{Retreat dynamics and the geometric lower bound}
\label{sec:lower_bound}
%======================================================================

\subsection{Setup}

Consider a marine-terminating glacier with grounding-line position $\xg(t)$ on a retrograde bed
\begin{equation}
	b(x) = -(b_0 + \alpha_b\, x),
	\label{eq:bed}
\end{equation}
where $b_0 > 0$ is the current bed depth and $\alpha_b > 0$ is the retrograde slope. At the grounding line, ice is at flotation:
\begin{equation}
	\hg(\xg) = \phi\,(b_0 + \alpha_b\, \xg) = h_0 + \phi\,\alpha_b\, \xg,
	\label{eq:flotation}
\end{equation}
with $\phi = \rhow/\rhoi \approx 1.12$ and $h_0 = \phi\, b_0$.

The grounding-line retreat rate is \citep{Schoof2012}
\begin{equation}
	\dot{\xg} = \frac{\qg - q_{\mathrm{in}}}{\hg},
	\label{eq:retreat_rate}
\end{equation}
where $q_{\mathrm{in}}$ is the upstream resupply flux. Setting $q_{\mathrm{in}} = 0$ and $\dot{a} = 0$ (zero accumulation) gives the fastest possible retreat. In this limit, with $\qg = q_0(\hg/h_0)^{s_{\mathrm{full}}}$ calibrated to the current discharge:
\begin{equation}
	\frac{\dd\hhat}{\dd\tau} = \hhat^{\,s_{\mathrm{full}}-1},
	\qquad \hhat(0) = 1,
	\label{eq:hhat_ode}
\end{equation}
where $\hhat = \hg/h_0$, $\tau = t/t_*$, and
\begin{equation}
	t_* = \frac{h_0}{\phi\,\alpha_b\,u_0}
	\label{eq:tstar}
\end{equation}
is the \emph{geometric timescale} ($u_0 = q_0/h_0$ is the current grounding-line velocity).

\subsection{Finite-time singularity}

For $s_{\mathrm{full}} > 2$ (all physically relevant cases, \cref{tab:exponents}):
\begin{equation}
	\hhat(\tau) = \bigl[1 - (s_{\mathrm{full}}-2)\,\tau\bigr]^{-1/(s_{\mathrm{full}}-2)},
	\label{eq:hhat_solution}
\end{equation}
which diverges at
\begin{equation}
	t_{\mathrm{sing}} = \frac{t_*}{s_{\mathrm{full}} - 2}.
	\label{eq:t_sing}
\end{equation}
For Thwaites Glacier (\cref{tab:parameters}; $t_* \approx 116$\,yr) with $n = 4$ ($s_{\mathrm{full}} = 5.8$): $t_{\mathrm{sing}} \approx 31$\,yr from onset, i.e., $\sim$2041. This is not a prediction of actual behavior---it reflects the extreme zero-resupply, zero-buttressing assumptions---but it demonstrates that the flux dynamics are \emph{fast} relative to the 21st-century timescale.

\subsection{The geometric mass-loss identity}

The total SLR from a marine ice sheet satisfies the mass conservation identity
\begin{equation}
	H_{\mathrm{SLR}}(T) = \frac{1}{\rhow\,\Aoc}\int_0^T\!\bigl(\Qg(t) - \dot{a}_{\mathrm{total}}\bigr)\,\dd t,
	\label{eq:slr_identity}
\end{equation}
where $\Qg = \rhoi\,\qg\,W$ is the total grounding-line discharge and $\dot{a}_{\mathrm{total}}$ is the total accumulation over the grounded domain. This identity holds for the full coupled system: sustaining grounding-line discharge $\Qg$ requires interior drawdown that reduces the mass above flotation. Even though ice at the grounding line is at flotation (zero local mass above flotation), the process of sustaining the discharge thins the interior, and it is this interior thinning that raises sea level.

In the zero-resupply, zero-accumulation limit, the total mass crossing the grounding line during the traverse of the retrograde section from $b_0$ to $b_{\max}$ can be computed by changing variables from $t$ to $\xg$. Since $\qg\,\dd t = \hg\,\dd\xg$ in this limit:
\begin{equation}
	\int_0^{t_{\max}}\!\!\qg\,\dd t
	= \int_0^{\Delta x}\!\hg\,\dd\xg
	= \int_0^{\Delta x}\!\phi(b_0 + \alpha_b\,x)\,\dd x
	= \frac{\phi}{2}\bigl(b_{\max}^2 - b_0^2\bigr)\,\frac{1}{\alpha_b}.
	\label{eq:traverse_integral}
\end{equation}

\begin{proposition}[Geometric lower bound]
\label{prop:lower_bound}
If MISI drives the grounding line through the full retrograde section (bed depths $b_0$ to $b_{\max}$), the resulting SLR satisfies
\begin{equation}
	\boxed{H_{\mathrm{SLR}} \geq \frac{\rhoi\,W\,\phi}{2\,\rhow\,\Aoc\,\alpha_b}\bigl(b_{\max}^2 - b_0^2\bigr) = \frac{Q_0\,t_*}{\rhow\,\Aoc}\cdot\frac{\hhat_{\max}^2 - 1}{2},}
	\label{eq:lower_bound}
\end{equation}
where $\hhat_{\max} = \phi\,b_{\max}/h_0$. This bound is independent of $n$, $m$, the rate factor $A$, and the sliding coefficient $C$. It depends only on the bed profile ($b_0$, $b_{\max}$, $\alpha_b$), glacier width ($W$), and the current state ($h_0$, $Q_0$).
\end{proposition}

\noindent The $s$-independence follows from the change of variables: $\qg\,\dd t = \hg\,\dd\xg$ eliminates the flux law entirely. All that $s$ controls is \emph{how fast} the traverse occurs, not \emph{how much mass} is lost during it.

\paragraph{Remark: why this is a lower bound.} The traverse integral counts only the discharge accumulated while the grounding line sweeps the retrograde section. After reaching maximum bed depth, the grounding line continues to discharge ice---potentially at very high rates---drawing down interior ice above flotation. This post-traverse contribution can be very large (see \cref{sec:upper_bound}). Moreover, in a fully coupled system with interior resupply, the total integrated discharge $\int \Qg\,\dd t$ exceeds the traverse integral even during the retrograde phase, because interior ice flowing toward the grounding line is an additional source of discharge mass. The traverse integral is therefore a strict lower bound on the SLR from MISI.

\subsection{Numerical evaluation}

\begin{table}[ht]
\centering
\caption{Representative parameters for Thwaites Glacier and the Amundsen Sea Embayment (ASE). Bed depths and slopes from BedMachine Antarctica \citep{Morlighem2020}. Discharges from \citet{Rignot2019}. The geometric timescale is $t_* = h_0/(\phi\,\alpha_b\,u_0)$.}
\label{tab:parameters}
\smallskip
\begin{tabular}{l c c l}
\toprule
Parameter & Thwaites & ASE (total) & Source \\
\midrule
GL width, $W$ & 120\,km & 300\,km & Remote sensing \\
Current bed depth, $b_0$ & 700\,m & 600--900\,m & BedMachine \\
Flotation thickness, $h_0$ & 785\,m & --- & Derived \\
Max trough depth, $b_{\max}$ & 2000\,m & 1200--2000\,m & BedMachine \\
$\hhat_{\max}$ & 2.85 & 1.5--2.9 & Derived \\
Retrograde slope, $\alpha_b$ & 0.0052 & 0.003--0.008 & BedMachine \\
Total discharge, $Q_0$ & 100\,Gt/yr & 250\,Gt/yr & \citet{Rignot2019} \\
GL velocity, $u_0$ & 1.2\,km/yr & 0.5--4\,km/yr & InSAR \\
$t_*$ & 116\,yr & 80--200\,yr & Eq.~\eqref{eq:tstar} \\
$\Vaf$ & $\sim$600\,mm & $\sim$1100\,mm & \citet{Morlighem2020} \\
\bottomrule
\end{tabular}
\end{table}

\Cref{tab:lower_bounds} evaluates the geometric lower bound for the major ASE glacier systems and for WAIS as a whole.

\begin{table}[ht]
\centering
\caption{Geometric lower bound on SLR from MISI retrograde retreat. These values are independent of $n$, $m$, and $s$. The traverse time $t_{\max}$ is shown for $n = 4$ ($s_{\mathrm{full}} = 5.8$) to confirm that the full retrograde section can be traversed within the century under zero-resupply dynamics.}
\label{tab:lower_bounds}
\smallskip
\begin{tabular}{l c c c}
\toprule
System & Lower bound (mm\,SLE) & $t_{\max}$ ($n{=}4$) & $\Vaf$ (mm\,SLE) \\
\midrule
Thwaites & 114 & $\sim$30\,yr & $\sim$600 \\
Pine Island & $\sim$30 & $\sim$20\,yr & $\sim$400 \\
Smith/Kohler & $\sim$10 & $\sim$15\,yr & $\sim$100 \\
\midrule
ASE total & $\sim$\textbf{155} & --- & $\sim$1100 \\
Other WAIS & $\sim$50 & --- & $\sim$2200 \\
\midrule
\textbf{Total WAIS} & $\sim$\textbf{200} & --- & $\sim$3300 \\
\bottomrule
\end{tabular}
\end{table}

The geometric lower bound for all of WAIS is approximately \textbf{200\,mm\,SLE}. This is the minimum SLR from MISI if the grounding lines traverse their full retrograde sections, regardless of the ice rheology. The traverse times ($\sim$15--30\,yr for individual ASE glaciers at $n = 4$) confirm that the full retrograde sections can be traversed well within the 21st century under the zero-resupply dynamics, so the bound is achievable within the projection horizon.


%======================================================================
\section{Buttressing-dependent upper bound}
\label{sec:upper_bound}
%======================================================================

\subsection{The zero-buttressing ceiling: $\Vaf$}

After the grounding line passes the deepest point of the retrograde section, discharge continues---potentially at very high rates. The peak flux ratio at maximum bed depth is $\hat{q}_{\max} = \hhat_{\max}^{s_{\mathrm{full}}}$. For Thwaites with $\hhat_{\max} = 2.85$ and $s_{\mathrm{full}} = 5.8$: $\hat{q}_{\max} \approx 390$. At 390 times the current discharge ($\sim$39{,}000\,Gt/yr), the entire Thwaites ice volume above flotation ($\sim$217{,}000\,Gt) would be drained in under 6 years.

This post-traverse flux is physically unrealizable at face value---it would require ice velocities $\sim$100$\times$ current values---but it establishes that in the zero-buttressing limit, the flux dynamics are fast enough to mobilize the \emph{entire} ice volume above flotation within the 21st century. The upper bound in this limit is simply
\begin{equation}
	H_{\mathrm{SLR}}^{\mathrm{max}} = \Vaf.
	\label{eq:Vaf_ceiling}
\end{equation}
For WAIS, $\Vaf \approx 3.3$\,m\,SLE---a trivially large bound that provides no useful constraint.

\subsection{Non-trivial bound with partial buttressing}

A useful upper bound requires residual buttressing ($\theta > 0$). With buttressing, the flux is reduced by a factor $(1 - \theta)^{s_b}$ relative to the unbuttressed case, \eqref{eq:buttressed_flux}. This has two effects:
\begin{enumerate}
	\item The \emph{instantaneous discharge rate} is reduced, slowing the drawdown of interior ice.
	\item The \emph{retreat dynamics} are slowed, because the flux-thickness feedback is weaker when buttressing absorbs part of the driving stress.
\end{enumerate}

For a constant buttressing ratio $\theta$, the retreat ODE becomes
\begin{equation}
	\frac{\dd\hhat}{\dd\tau} = (1-\theta)^{s_b}\,\hhat^{\,s_{\mathrm{full}}-1},
	\qquad \hhat(0) = 1,
	\label{eq:hhat_buttressed}
\end{equation}
which has the same analytical solution as \eqref{eq:hhat_solution} but with an effective timescale $t_*^{\mathrm{eff}} = t_*/(1-\theta)^{s_b}$. The singularity time becomes
\begin{equation}
	t_{\mathrm{sing}}^{\mathrm{eff}} = \frac{t_*}{(s_{\mathrm{full}} - 2)\,(1-\theta)^{s_b}}.
	\label{eq:tsing_buttressed}
\end{equation}
Even modest buttressing dramatically extends the singularity time. For Thwaites ($t_* = 116$\,yr, $s_{\mathrm{full}} = 5.8$) with Pegler buttressing ($s_b^{(P)} = 4.0$, appropriate for confined ASE glaciers):

\begin{table}[ht]
\centering
\caption{Effect of buttressing on Thwaites dynamics ($n = 4$, Pegler framework). $\theta$: buttressing ratio. $t_{\mathrm{sing}}^{\mathrm{eff}}$: time to mathematical singularity. $\hat{q}(T)$: discharge amplification at 2100. $M(T)$: cumulative discharge (mm\,SLE), capped at $\Vaf = 600$\,mm.}
\label{tab:buttressing}
\smallskip
\begin{tabular}{c c c c l}
\toprule
$\theta$ & $(1-\theta)^{s_b}$ & $t_{\mathrm{sing}}^{\mathrm{eff}}$ (yr) & $\hat{q}(T=90)$ & $M$ (mm\,SLE) \\
\midrule
0    & 1.000 & 31   & $\gg 1$ & 600 (= $\Vaf$) \\
0.1  & 0.656 & 47   & $\gg 1$ & 600 (= $\Vaf$) \\
0.2  & 0.410 & 75   & $\gg 1$ & 600 (= $\Vaf$) \\
0.3  & 0.240 & 127  & 129 & $\sim$570 \\
0.4  & 0.130 & 235  & 10.6 & $\sim$300 \\
0.5  & 0.063 & 490  & 3.1 & $\sim$160 \\
0.6  & 0.026 & 1180 & 1.6 & $\sim$100 \\
0.7  & 0.008 & 3700 & 1.1 & $\sim$70 \\
\bottomrule
\end{tabular}
\end{table}

The table reveals a sharp transition: for $\theta \lesssim 0.3$, buttressing is insufficient to prevent the singularity from being reached within the century, and the upper bound saturates at $\Vaf$. For $\theta \gtrsim 0.4$, the dynamics are slow enough that only a fraction of $\Vaf$ is mobilized, yielding a non-trivial upper bound.

The cumulative discharge at year $T$, accounting for the bed-geometry truncation, is
\begin{equation}
	M(T) = \min\!\left(\Vaf,\;\; Q_0\,t_*^{\mathrm{eff}}\left[\frac{\hhat(T)^2 - 1}{2} + \hhat_{\max}^{s_{\mathrm{full}}}\cdot\bigl(\That^{\mathrm{eff}} - \tau_{\max}\bigr)^+\right]\right),
	\label{eq:M_buttressed}
\end{equation}
where $\That^{\mathrm{eff}} = T/t_*^{\mathrm{eff}}$ and $\hhat(T)$ is evaluated from \eqref{eq:hhat_solution} at $\tau = T/t_*^{\mathrm{eff}}$ (or $\hhat_{\max}$ if the traverse completes before $T$).

\subsection{Buttressing loss trajectory}

In reality, $\theta$ is not constant: buttressing decreases as ice shelves thin and retreat. A time-varying $\theta(t)$ can be incorporated by solving \eqref{eq:hhat_buttressed} numerically, or by specifying a buttressing-loss timescale $\tau_\theta$ and parameterizing $\theta(t) = \theta_0\,e^{-t/\tau_\theta}$ (exponential loss from initial value $\theta_0$). For rapid shelf collapse ($\tau_\theta \sim 10$\,yr), the system transitions quickly from the buttressed to the unbuttressed regime, and the upper bound approaches $\Vaf$. For gradual thinning ($\tau_\theta \sim 100$\,yr), significant buttressing persists throughout the century, keeping the upper bound well below $\Vaf$.

\citet{Naughten2023} showed that West Antarctic ice-shelf basal melt rates of $\sim$3$\times$ historical are unavoidable regardless of emissions scenario, implying $\tau_\theta \sim 50$--100\,yr for melt-driven buttressing loss. \citet{Joughin2014} documented ongoing retreat of the Thwaites grounding line on a retrograde bed, suggesting the transition from buttressed to partially unbuttressed dynamics is already underway.


%======================================================================
\section{Connection to the A4 scenario framework}
\label{sec:a4_connection}
%======================================================================

\subsection{Mapping scenarios to buttressing states}

The A4 scenario framework defines three WAIS regimes: S1 (status quo, 25--85\,mm), S2 (MISI, 120--1000\,mm), and S3 (MISI + MICI, 600--1400\,mm). The bounds derived here provide physical interpretation for these ranges:

\paragraph{S1 (25--85\,mm).} No MISI: the grounding line does not traverse the retrograde section. Discharge responds linearly to ocean thermal forcing without instability amplification, with bounds derived from the Joughin et al.\ (2021) melt--discharge linearity and projected Amundsen shelf warming. This corresponds to $\theta$ remaining large enough to stabilize the grounding line.

\paragraph{S2 lower bound ($\sim$120\,mm).} Represents late-onset MISI with high residual buttressing, limited primarily to the Thwaites basin. The value is anchored by rate separation from S1 (the instability feedback must produce contributions clearly above the 85\,mm S1 upper bound) and by bed geometry (even partial traverse of Thwaites' retrograde section produces $\sim$50--60\,mm\,SLE, which combined with melt-driven contributions from other basins exceeds 120\,mm). The geometric lower bound for the full ASE ($\sim$155\,mm, \cref{tab:lower_bounds}) lies above the S2 lower bound, consistent with the interpretation that 120\,mm represents early or partial MISI that has not yet driven grounding lines through their complete retrograde sections.

\paragraph{S2 range (120--1000\,mm).} The full S2 range corresponds to buttressing ratios of $\theta \approx 0.7$ (lower end, $\sim$70\,mm for Thwaites) to $\theta \approx 0.2$ (upper end, approaching $\Vaf$ for individual glaciers). The S2 upper bound of 1000\,mm requires either low residual buttressing ($\theta \lesssim 0.3$) across most ASE basins, or amplifying feedbacks (calving, subglacial hydrology \citep{Dow2022}) that effectively reduce $\theta$ faster. The positive skewness ($\alpha = +4$) assigned to S2 is consistent with the sharp nonlinearity in \cref{tab:buttressing}: a modest reduction in buttressing (from $\theta = 0.5$ to $\theta = 0.3$) more than triples the mass loss.

\paragraph{S3 (600--1400\,mm).} Exceeds the Thwaites $\Vaf$ of $\sim$600\,mm and is anchored at the upper end to the ASE $\Vaf$ of $\sim$1100\,mm with rheology correction ($\times\,1.28$, giving $\sim$1400\,mm). Reaching these values requires near-complete deglaciation of ASE basins, potentially aided by marine ice cliff instability (MICI) \citep{Bassis2012, DeContoP2016}. Recent evidence suggests MICI is unlikely to dominate during the 21st century \citep{Morlighem2024, Clerc2019, Schlemm2022}, consistent with the low weight ($P = 0.10$) and negative skewness ($\alpha = -3$) assigned to S3.

\subsection{The trajectory exponent $\beta$}

The A4 framework parameterizes temporal structure through $H(t) = H_{2100}\,(t/T)^\beta$. The connection to buttressing is:

\begin{itemize}
	\item High $\theta$ (strong buttressing) $\Rightarrow$ slow retreat, nearly constant discharge $\Rightarrow$ $\beta \approx 1$ (linear).
	\item Low $\theta$ (weak buttressing) $\Rightarrow$ fast retreat, accelerating discharge $\Rightarrow$ $\beta > 1$ (back-loaded).
	\item Decreasing $\theta(t)$ (progressive shelf loss) $\Rightarrow$ increasing flux nonlinearity through the century $\Rightarrow$ strongly back-loaded ($\beta \gg 1$).
\end{itemize}

The A4 prior on $\beta$ for S2 ($\mathrm{LN}(\ln 1.8,\, 0.3)$, median 1.8) is consistent with gradual buttressing loss over the century, transitioning from a buttressed regime ($\beta \approx 1$) in the early decades to a weakly buttressed regime ($\beta > 2$) later.

The buttressing exponents $s_b$ enter through the rheology correction: when correcting from $n_{\mathrm{ref}} = 3$ to $n = 4$, the trajectory exponent scales as $\beta(n)/\beta(n_{\mathrm{ref}}) = s_b(n)/s_b(n_{\mathrm{ref}})$. For the Pegler framework: $\beta(n=4)/\beta(n=3) = 4.0/3.0 = 1.33$; correcting $\beta = 1.8$ at $n=3$ gives $\beta \approx 2.4$ at $n = 4$.


%======================================================================
\section{Discussion}
\label{sec:discussion}
%======================================================================

\subsection{Complementarity of the two bounds}

The geometric lower bound and the buttressing-dependent upper bound bracket the MISI contribution from opposite directions:

\begin{itemize}
	\item The \textbf{lower bound} ($\sim$200\,mm) answers: \emph{what is the minimum SLR from MISI?} It is rheology-independent and depends only on what we know well (bed geometry, current discharge). Its weakness is that it only counts mass from the retrograde traverse, ignoring the potentially larger contribution from interior drawdown.

	\item The \textbf{upper bound} ($\Vaf$ at zero buttressing; non-trivial with partial buttressing) answers: \emph{what is the maximum SLR given residual buttressing?} It accounts for the full interior ice reservoir but depends on the buttressing ratio $\theta$ and the framework-specific exponent $s_b$, both of which are uncertain.
\end{itemize}

Together, they establish that the S2 range (120--1000\,mm) is physically well-motivated: the lower end represents late-onset, partial MISI (rate separation from S1 and partial retrograde traverse), the geometric lower bound ($\sim$200\,mm) sits within the lower quartile of S2, and the upper end is set by the rate of buttressing loss (requiring near-complete shelf collapse across most ASE basins).

\subsection{Comparison with Pfeffer et al.\ (2008)}

\citet{Pfeffer2008} bounded glacier contributions to SLR by prescribing velocity amplification factors. Our approach derives the amplification from the flux law, distinguishing between the full thickness exponent $s_{\mathrm{full}}$ (which controls retreat speed) and the buttressing exponent $s_b$ (which controls sensitivity to shelf loss). The geometric lower bound is more robust than Pfeffer's kinematic bound because it is independent of the assumed velocity amplification.

\subsection{Limitations}

\begin{enumerate}
	\item \textbf{Linear bed.} Real beds have local ridges and variable slope \citep{Morlighem2020}. The lower bound should be refined by integrating over actual BedMachine profiles: $M_{\mathrm{traverse}} = \rhoi W \int_{x_0}^{x_{\max}} \phi\, |b(x)|\,\dd x$.

	\item \textbf{Constant $\theta$.} Buttressing evolves with ice-shelf geometry. A realistic upper bound requires coupling $\theta(t)$ to an ice-shelf evolution model, or parameterizing the buttressing-loss trajectory.

	\item \textbf{Width variation.} The Thwaites trough broadens inland from $\sim$120\,km to $\sim$200\,km. Width variation enters the lower bound through $W(x)$ in the traverse integral.

	\item \textbf{Steady-state flux law.} During rapid retreat, transient effects may cause the flux to deviate from the steady-state boundary-layer scaling \citep{Schoof2012, HaseloffSergienko2022}. The steady-state flux law approximates the transient behavior well when the retreat rate is small compared to the ice velocity, a condition that breaks down near the singularity.
\end{enumerate}


%======================================================================
\section*{Summary}
%======================================================================

The grounding-line flux law provides two complementary bounds on 21st-century WAIS mass loss from MISI:

\begin{enumerate}
	\item A \textbf{geometric lower bound} of $\sim$200\,mm\,SLE, equal to the mass discharged during grounding-line retreat through the full retrograde bed sections of all WAIS basins. This bound is independent of the Glen's law exponent $n$, the sliding law exponent $m$, the rate factor $A$, and the sliding coefficient $C$. It depends only on the bed geometry, glacier width, and current discharge---quantities known from BedMachine and satellite observations. If MISI is triggered, this is the minimum guaranteed SLR.

	\item A \textbf{buttressing-dependent upper bound} that transitions from $\sim$100\,mm (60--70\% buttressing retained) through $\sim$300\,mm ($\sim$40\% retained) to the full ice volume $\Vaf \approx 3.3$\,m ($< 30\%$ retained). The transition is sharp: a reduction in buttressing from 50\% to 30\% more than triples the projected mass loss, reflecting the high buttressing exponents ($s_b \sim 3$--$4$) predicted by the Pegler and Haseloff frameworks.
\end{enumerate}

The A4 S2 scenario range (120--1000\,mm) spans from late-onset partial MISI (below the geometric lower bound of $\sim$200\,mm, representing early or incomplete retrograde traverse) to the regime of substantial interior drawdown under low buttressing. The upper end requires near-complete shelf loss across most ASE basins. Both ends are informed by the flux-law physics, applied at opposite ends of the buttressing spectrum.


%======================================================================
% References
%======================================================================
\bibliographystyle{agufull08}

\begin{thebibliography}{99}

\bibitem[Bassis and Walker(2012)]{Bassis2012}
Bassis, J.~N. and Walker, C.~C. (2012).
Upper and lower limits on the stability of calving glaciers from the yield strength envelope of ice.
\textit{Proceedings of the Royal Society A}, 468(2140), 913--931.
\doi{10.1098/rspa.2011.0422}

\bibitem[Clerc et~al.(2019)]{Clerc2019}
Clerc, F., Minchew, B.~M., and Behn, M.~D. (2019).
Marine ice cliff instability mitigated by slow removal of ice shelves.
\textit{Geophysical Research Letters}, 46(21), 12108--12116.
\doi{10.1029/2019GL084183}

\bibitem[DeConto and Pollard(2016)]{DeContoP2016}
DeConto, R.~M. and Pollard, D. (2016).
Contribution of {A}ntarctica to past and future sea-level rise.
\textit{Nature}, 531(7596), 591--597.
\doi{10.1038/nature17145}

\bibitem[Dow et~al.(2022)]{Dow2022}
Dow, C.~F., Ross, N., Jeofry, H., Siu, K., and Siegert, M.~J. (2022).
Antarctic basal environment shaped by high-pressure flow through a subglacial river system.
\textit{Nature Geoscience}, 15, 892--898.
\doi{10.1038/s41561-022-01059-1}

\bibitem[Getraer and Morlighem(2025)]{GetraerMorlighem2025}
Getraer, G. and Morlighem, M. (2025).
The impact of a higher-order creep law on {A}ntarctic ice sheet projections.
\textit{Geophysical Research Letters}, 52.
\doi{10.1029/2024GL112516}

\bibitem[Goelzer et~al.(2020)]{Goelzer2020}
Goelzer, H., Nowicki, S., Payne, A., et~al. (2020).
The future sea-level contribution of the {G}reenland ice sheet: a multi-model ensemble study of {ISMIP6}.
\textit{The Cryosphere}, 14(9), 3071--3096.
\doi{10.5194/tc-14-3071-2020}

\bibitem[Haseloff and Sergienko(2018)]{HaseloffSergienko2018}
Haseloff, M. and Sergienko, O.~V. (2018).
The effect of buttressing on grounding line dynamics.
\textit{Journal of Glaciology}, 64(245), 417--431.
\doi{10.1017/jog.2018.30}

\bibitem[Haseloff and Sergienko(2022)]{HaseloffSergienko2022}
Haseloff, M. and Sergienko, O.~V. (2022).
Effects of calving and submarine melting on steady states and stability of buttressed marine ice sheets.
\textit{Journal of Glaciology}, 68(272), 1149--1166.
\doi{10.1017/jog.2022.29}

\bibitem[Joughin et~al.(2014)]{Joughin2014}
Joughin, I., Smith, B.~E., and Medley, B. (2014).
Marine ice sheet collapse potentially under way for the {T}hwaites {G}lacier {B}asin, {W}est {A}ntarctica.
\textit{Science}, 344(6185), 735--738.
\doi{10.1126/science.1249055}

\bibitem[Martin et~al.(2026)]{Martin2026}
Martin, D.~F., Trevers, R., and Minchew, B.~M. (2026).
A higher-order flow law produces systematic under-estimation of sea-level rise.
\textit{AGU Advances}, 7.
\doi{10.1029/2025AV001946}

\bibitem[Millstein et~al.(2022)]{Millstein2022}
Millstein, J.~D., Minchew, B.~M., and Pegler, S.~S. (2022).
Ice viscosity is more sensitive to stress than commonly assumed.
\textit{Communications Earth \& Environment}, 3, 158.
\doi{10.1038/s43247-022-00485-8}

\bibitem[Morlighem et~al.(2020)]{Morlighem2020}
Morlighem, M., Rignot, E., Binder, T., et~al. (2020).
Deep glacial troughs and stabilizing ridges unveiled beneath the margins of the {A}ntarctic ice sheet.
\textit{Nature Geoscience}, 13, 132--137.
\doi{10.1038/s41561-019-0510-8}

\bibitem[Morlighem et~al.(2024)]{Morlighem2024}
Morlighem, M., Goldberg, D., Barnes, J.~M., et~al. (2024).
The {W}est {A}ntarctic {I}ce {S}heet may not be vulnerable to marine ice cliff instability during the 21st century.
\textit{Science Advances}, 10, eado7794.
\doi{10.1126/sciadv.ado7794}

\bibitem[Naughten et~al.(2023)]{Naughten2023}
Naughten, K.~A., Holland, P.~R., and De~Rydt, J. (2023).
Unavoidable future increase in {W}est {A}ntarctic ice-shelf melting over the twenty-first century.
\textit{Nature Climate Change}, 13, 1222--1228.
\doi{10.1038/s41558-023-01818-x}

\bibitem[Pegler(2018)]{Pegler2018}
Pegler, S.~S. (2018).
Suppression of marine ice sheet instability.
\textit{Journal of Fluid Mechanics}, 857, 648--680.
\doi{10.1017/jfm.2018.776}

\bibitem[Pfeffer et~al.(2008)]{Pfeffer2008}
Pfeffer, W.~T., Harper, J.~T., and O'Neel, S. (2008).
Kinematic constraints on glacier contributions to 21st-century sea-level rise.
\textit{Science}, 321(5894), 1340--1343.
\doi{10.1126/science.1159099}

\bibitem[Rignot et~al.(2019)]{Rignot2019}
Rignot, E., Mouginot, J., Scheuchl, B., van~den~Broeke, M., van~Wessem, M.~J., and Morlighem, M. (2019).
Four decades of {A}ntarctic {I}ce {S}heet mass balance from 1979--2017.
\textit{Proceedings of the National Academy of Sciences}, 116(4), 1095--1103.
\doi{10.1073/pnas.1812883116}

\bibitem[Robel et~al.(2019)]{Robel2019}
Robel, A.~A., Seroussi, H., and Roe, G.~H. (2019).
Marine ice sheet instability amplifies and skews uncertainty in projections of future sea-level rise.
\textit{Proceedings of the National Academy of Sciences}, 116(30), 14887--14892.
\doi{10.1073/pnas.1904822116}

\bibitem[Schlemm et~al.(2022)]{Schlemm2022}
Schlemm, T., Feldmann, J., Winkelmann, R., and Levermann, A. (2022).
Stabilizing effect of m\'{e}lange buttressing on the marine ice-cliff instability of the {W}est {A}ntarctic {I}ce {S}heet.
\textit{The Cryosphere}, 16, 1979--1996.
\doi{10.5194/tc-16-1979-2022}

\bibitem[Schoof(2007)]{Schoof2007}
Schoof, C. (2007).
Ice sheet grounding line dynamics: Steady states, stability, and hysteresis.
\textit{Journal of Geophysical Research: Earth Surface}, 112(F3), F03S28.
\doi{10.1029/2006JF000664}

\bibitem[Schoof(2012)]{Schoof2012}
Schoof, C. (2012).
Marine ice sheet stability.
\textit{Journal of Fluid Mechanics}, 698, 62--72.
\doi{10.1017/jfm.2012.43}

\end{thebibliography}

\end{document}
